---
title: "DAT 490 Capstone: Data Cleaning and Spatial Join"
output:
  pdf_document:
    toc: true
    toc_depth: 2
    number_sections: true
    latex_engine: xelatex
---

\newpage

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
options(tigris_use_cache = TRUE)
```

```{r libraries, include=FALSE}
library(tidyverse)
library(lubridate)
library(janitor)
library(skimr)
library(here)
library(tidycensus)
library(sf)
library(tigris)
library(kableExtra)
library(mice)
```

# LAPD Data

## Load Raw Data

```{r lapd-load, echo = FALSE, warning = FALSE}
lapd_raw <- read_csv(here("Crime_Data_from_2020_to_2024.csv"),
                     show_col_types = FALSE) |>
  clean_names()

head(lapd_raw)
```
 
## Select Columns

```{r lapd-select-cols}
lapd <- lapd_raw |>
  select(
    rpt_no       = dr_no,        ##report number
    date_occ,
    time_occ,
    crime_cd     = crm_cd,       ##crime code
    case_status  = status_desc,  ##case status
    lat,
    lon
  )

head(lapd)
```

## Clean Time Data

```{r lapd-dates}
lapd <- lapd |>
  mutate(
    date_occ = as.Date(mdy_hms(date_occ)),
    year_occ = year(date_occ)
  )
```

## Clean Coordinates & Flag Invalid Locations

```{r lapd-coordinates}
lapd <- lapd |>
  mutate(
    lat_clean = if_else(
      lat == 0 | lon == 0 |
        (round(lat, 4) == 34.0195 & round(lon, 4) == -118.4108),
      NA_real_, lat
    ),
    lon_clean = if_else(
      lat == 0 | lon == 0 |
        (round(lat, 4) == 34.0195 & round(lon, 4) == -118.4108),
      NA_real_, lon
    ),
    has_coords = !is.na(lat_clean) & !is.na(lon_clean)
  )

cat("Records with valid coordinates:", sum(lapd$has_coords), "\n", 
    "Records with missing coordinates removed:  ", sum(!lapd$has_coords), "\n", 
    sep = "")

lapd <- lapd |> filter(has_coords)

```

## Remove Duplicates

```{r lapd-dup}
n_before <- nrow(lapd)

lapd <- lapd |>
  distinct(rpt_no, .keep_all = TRUE)

cat("Duplicate rows removed:", n_before - nrow(lapd), "\n",
    "Final row count:       ", nrow(lapd), "\n",
    sep = "")
```

## Missing Values Check

```{r lapd-missing-values}
n_missing <- sum(is.na(lapd))
cat("Total missing values in cleaned LAPD data: ", n_missing, "\n")
```

\newpage

## Derive Variables

```{r lapd-derive}
lapd <- lapd |>
  mutate(
    ##crime type categories based on ucr/compstat codes
    crime_category = case_when(
      crime_cd %in% c(110, 113)                            ~ "Homicide",
      crime_cd %in% c(210, 220)                            ~ "Robbery",
      crime_cd %in% c(230, 231, 235, 236, 250, 251,
                      761, 926)                            ~ "Assault",
      crime_cd %in% c(624, 625, 626, 627, 628, 629)        ~ "Simple Assault",
      crime_cd %in% c(121, 122, 815, 820, 821,
                      822, 860)                            ~ "Sexual Assault",
      crime_cd %in% c(760, 762, 810, 812, 813,
                      814, 850)                            ~ "Sex Offenses",
      crime_cd %in% c(310, 320)                            ~ "Burglary",
      crime_cd %in% c(330, 331, 341, 345, 354, 410, 420,
                      421, 440, 441, 444, 445, 450, 451,
                      452, 453, 470, 471, 473, 474, 475,
                      480, 485, 487, 950)                  ~ "Theft",
      crime_cd %in% c(510, 520, 433)                       ~ "Vehicle Theft",
      crime_cd %in% c(343, 350, 351, 352, 353,
                      442, 443)                            ~ "Shoplifting",
      crime_cd %in% c(740, 745)                            ~ "Vandalism",
      crime_cd %in% c(865, 940)                            ~ "Drug/Narcotic",
      crime_cd %in% c(753, 756)                            ~ "Weapons",
      crime_cd %in% c(648)                                 ~ "Arson",
      TRUE                                                 ~ "Other"
    ),
    ##violent crime flag
    is_violent = crime_category %in% c(
      "Homicide", "Robbery", "Assault", "Sexual Assault",
      "Sex Offenses", "Weapons", "Arson"
    ),

    ##property crime flag
    is_property = crime_category %in% c(
      "Burglary", "Theft", "Vehicle Theft", "Shoplifting", "Vandalism"
    ),

    ##binary outcome for rq1: violent (1) vs property (0)
    ##incidents classified as neither are excluded from rq1 modeling
    crime_type = case_when(
      is_violent  ~ "Violent",
      is_property ~ "Property",
      TRUE        ~ "Other"
    )
  )
```

## Export Clean LAPD Data
```{r lapd-export}
if (!dir.exists("output")) dir.create("output")

if (!file.exists("output/lapd_clean.rds")) {
  saveRDS(lapd, "output/lapd_clean.rds")
  write_csv(lapd, "output/lapd_clean.csv")
  cat("Files saved to /output:\n",
      "                         lapd_clean.rds\n",
      "                         lapd_clean.csv\n",
      sep = "")
} else {
  cat("Output files already exist.\n")
}
```

\newpage 

# ACS Census Data

## Load Raw Data

```{r acs-load, message=FALSE}
acs_vars <- c(
  ##economic vulnerability
  median_income        = "B19013_001",
  poverty_pop          = "B17001_001",  ##population (excluding group quarters)
  poverty              = "B17001_002",  ##population below poverty line
  labor_total          = "B23025_003",  ##labor force
  labor_unemployed     = "B23025_005",
  gini                 = "B19083_001",  ##measure of income inequality within tract
  housing_burden_rent  = "B25070_010",  ##renters spending 50%+ on housing
  housing_burden_mort  = "B25091_011",  ##mortgage holders spending 50%+ on housing
  no_vehicle_renter    = "B25044_003",
  no_vehicle_owner     = "B25044_010",

  ##neighborhood stability
  total_housing        = "B25002_001",  
  vacant_units         = "B25002_003",  
  occupied_units       = "B25003_001",  
  renter_units         = "B25003_003",  ##renter-occupied units vs owner-occupied
  building_units       = "B25024_001",  ##number of units in building

  ##social disorganization
  single_parent_f      = "B11012_010",  ##single mother households
  single_parent_m      = "B11012_015",  ##single father households
  lang_iso_total       = "C16002_001",  ##total households limited english
  lang_iso_spanish     = "C16002_004",  ##spanish limited english
  lang_iso_indoeur     = "C16002_007",  ##other indo-european limited english
  lang_iso_asian       = "C16002_010",  ##asian/pacific island limited english
  lang_iso_other       = "C16002_013",  ##other languages limited english
  internet_access      = "B28002_002",  ##households with internet
  
  ##demographics
  total_pop            = "B01003_001",
  median_age           = "B01002_001",
  household_size       = "B25010_001",
  white                = "B03002_003",  
  black                = "B03002_004",
  native               = "B03002_005",
  asian                = "B03002_006",
  pacific_islander     = "B03002_007",  ##native hawaiian and other pacific islander
  other_race           = "B03002_008",
  two_or_more          = "B03002_009",  ##two or more races
  hispanic             = "B03002_012",  ##mutually exclusive with race categories
  edu_total            = "B15003_001",  ##total population 25 and over
  edu_no_hs            = "B15003_002",  ##no schooling completed
  edu_9th              = "B15003_013",  
  edu_10th             = "B15003_014",  
  edu_11th             = "B15003_015",  
  edu_12th             = "B15003_016",   
  edu_hs_diploma       = "B15003_017",
  edu_ged              = "B15003_018",  ##GED or alternative credential
  edu_some_college     = "B15003_019",  ##some college less than 1 year
  edu_some_college_m   = "B15003_020",  ##some college 1 or more years no degree
  edu_associates       = "B15003_021",
  edu_bachelors        = "B15003_022",
  edu_masters          = "B15003_023",
  edu_professional     = "B15003_024",  ##professional school degree
  edu_doctorate        = "B15003_025",
  disability_min       = "B18135_003",  ##under 19 with a disability
  disability_mid       = "B18135_014",  ##19-64 with a disability
  disability_eld       = "B18135_025",  ##65+ with a disability
  disability_pop       = "B18135_001"   ##total noninstitutionalized population
)

##only pull from API if raw file doesn't already exist
if (!file.exists("output/acs_raw.rds")) {
  census_api_key(Sys.getenv("CENSUS_API_KEY"))
  acs_raw <- get_acs(
    geography = "tract",
    variables = acs_vars,
    state     = "CA",
    county    = "Los Angeles",
    year      = 2024,
    survey    = "acs5",
    geometry  = TRUE
  )
  saveRDS(acs_raw, "output/acs_raw.rds")
  cat("ACS data downloaded and saved.\n")
} else {
  acs_raw <- readRDS("output/acs_raw.rds")
  cat("ACS data loaded from cache.\n")
}

head(acs_raw)
```

## Flag Unreliable Estimates

**Note:** Several variables show high rates of unreliable estimates across census tracts, particularly small subgroup counts. This is expected in small-population tracts, but should be interpreted with caution.

```{r acs-moe}
##flag variables where margin of error exceeds 50% of the estimate
moe_check <- acs_raw |>
  st_drop_geometry() |>
  filter(!is.na(estimate), estimate != 0) |>
  mutate(moe_pct = moe / estimate) |>
  filter(moe_pct > 0.5) |>
  count(variable, name = "n_unreliable") |>
  arrange(desc(n_unreliable))

cat("Variables with unreliable estimates (MOE > 50%):\n")
knitr::kable(moe_check, col.names = c("Variable", "Tracts with Unreliable Estimates"))
```

## Reshape Data

```{r acs-reshape}
##pivot acs data to wide so each tract is one row
acs <- acs_raw |>
  select(GEOID, NAME, variable, estimate, geometry) |>
  pivot_wider(names_from = variable, values_from = estimate)
```

## Remove Zero Population Tracts

```{r acs-zero-pop}
n_before <- nrow(acs)

acs <- acs |>
  filter(!is.na(total_pop) & total_pop > 0)

cat("Zero/null population tracts removed: ", n_before - nrow(acs), "\n",
    "Remaining tracts:                    ", nrow(acs), "\n",
    sep = "")
```

## Derive Variables

```{r acs-derive}
acs <- acs |>
  mutate(
    ##economic vulnerability
    poverty_rate         = poverty / poverty_pop,
    unemployment_rate    = labor_unemployed / labor_total,
    no_vehicle_rate      = (no_vehicle_renter + no_vehicle_owner) / occupied_units,
    housing_burden_rate  = ((housing_burden_rent + housing_burden_mort) / occupied_units),

    ##neighborhood stability
    vacancy_rate         = vacant_units / total_housing,
    renter_rate          = renter_units / occupied_units,

    ##social disorganization
    single_parent_rate   = (single_parent_f + single_parent_m) / occupied_units,
    lang_isolation_rate  = (lang_iso_spanish + lang_iso_indoeur +
                            lang_iso_asian + lang_iso_other) / lang_iso_total,
    internet_rate        = internet_access / occupied_units,

    ##demographics
    pct_white            = white / total_pop,
    pct_black            = black / total_pop,
    pct_native           = native / total_pop,
    pct_asian            = asian / total_pop,
    pct_pacific          = pacific_islander / total_pop,
    pct_other            = other_race / total_pop,
    pct_two_or_more      = two_or_more / total_pop,
    pct_hispanic         = hispanic / total_pop,
    edu_rate_no_hsdip    = (edu_no_hs + edu_9th + edu_10th + edu_11th + 
                            edu_12th) / edu_total,
    edu_rate_hs          = (edu_hs_diploma + edu_ged) / edu_total,
    edu_rate_some_col    = (edu_some_college + edu_some_college_m + 
                            edu_associates) / edu_total,
    edu_rate_college     = (edu_bachelors + edu_masters + edu_professional + 
                            edu_doctorate) / edu_total,
    disability_rate      = (disability_min + disability_mid + 
                            disability_eld) / disability_pop
  )
```

## Check Missing Values

```{r acs-missing, echo = FALSE}
acs |>
  st_drop_geometry() |>
  summarise(across(everything(), \(x) sum(is.na(x)))) |>
  pivot_longer(everything(), names_to = "variable", values_to = "n_missing") |>
  filter(n_missing > 0) |>
  arrange(desc(n_missing)) |>
  knitr::kable(col.names = c("Variable", "Missing Values"))
```

## Impute Missing Values

```{r m-imputation, warning = FALSE}
##multiple imputation for missing acs predictors
acs_imp <- acs |>
  st_drop_geometry() |>
  select(GEOID, median_income, gini, no_vehicle_rate, housing_burden_rate, 
         vacancy_rate, renter_rate, single_parent_rate, lang_isolation_rate, 
         internet_rate, poverty_rate, unemployment_rate, disability_rate, 
         median_age, edu_rate_no_hsdip, edu_rate_college) |>
  mice(m = 5, method = "pmm", seed = 094, printFlag = FALSE)

imp_check <- lapply(1:5, function(i) {
  complete(acs_imp, i) |>
    filter(is.na(acs |> st_drop_geometry() |> pull(median_income)) |
           is.na(acs |> st_drop_geometry() |> pull(gini))) |>
    summarise(
      dataset       = i,
      median_income_mean = mean(median_income, na.rm = TRUE),
      median_income_sd   = sd(median_income,   na.rm = TRUE),
      gini_mean          = mean(gini,          na.rm = TRUE),
      gini_sd            = sd(gini,            na.rm = TRUE)
    )
}) |>
  bind_rows()

knitr::kable(imp_check, digits = 4,
             col.names = c("Dataset", "Income Mean", "Income SD",
                           "Gini Mean", "Gini SD"))

acs_imputed <- complete(acs_imp, 1) |>
  select(GEOID, median_income, gini, no_vehicle_rate, housing_burden_rate, 
         vacancy_rate, renter_rate, single_parent_rate, lang_isolation_rate, 
         internet_rate, poverty_rate, unemployment_rate, disability_rate, 
         median_age, edu_rate_no_hsdip, edu_rate_college)
acs <- acs |>
  select(-any_of(names(acs_imputed)[names(acs_imputed) != "GEOID"])) |>
  left_join(acs_imputed, by = "GEOID")

cat("Missing values after imputation: ",
    sum(is.na(acs |> st_drop_geometry() |>
                select(median_income, gini, no_vehicle_rate, housing_burden_rate, 
                       vacancy_rate, renter_rate, single_parent_rate, 
                       lang_isolation_rate, internet_rate, poverty_rate))), "\n")
```    
   
\newpage

## Drop Raw Input Variables

```{r acs-cleanup}
##drop raw acs input variables used only to compute derived rates
acs_raw_vars_drop <- c(
  ##economic vulnerability inputs
  "poverty", "poverty_pop", "labor_total", "labor_unemployed","housing_burden_rent",
  "housing_burden_mort", "no_vehicle_renter", "no_vehicle_owner",

  ##neighborhood stability inputs
  "total_housing", "vacant_units", "occupied_units", "renter_units",

  ##social disorganization inputs
  "single_parent_f", "single_parent_m", "lang_iso_total", "lang_iso_spanish",
  "lang_iso_indoeur", "lang_iso_asian", "lang_iso_other", "internet_access",

  ##demographics inputs
  "white", "black", "native", "asian", "pacific_islander",
  "other_race", "two_or_more", "hispanic",

  ##education inputs
  "edu_total", "edu_no_hs", "edu_9th", "edu_10th", "edu_11th",
  "edu_12th", "edu_hs_diploma", "edu_ged", "edu_some_college",
  "edu_some_college_m", "edu_associates", "edu_bachelors",
  "edu_masters", "edu_professional", "edu_doctorate",

  ##disability inputs
  "disability_min", "disability_mid", "disability_eld", "disability_pop"
)

acs <- acs |>
  select(-any_of(acs_raw_vars_drop))

cat("Raw input variables dropped: ", length(acs_raw_vars_drop), "\n",
    "Remaining ACS columns:       ", ncol(acs) - 2, "\n", 
    sep = "")
```

## Export Clean ACS Data

```{r acs-export, echo = FALSE}
if (!file.exists("output/acs_clean.rds")) {
  saveRDS(acs, "output/acs_clean.rds")
  acs |>
    st_drop_geometry() |>
    write_csv("output/acs_clean.csv")
  cat("Files saved to /output:\n",
      "                        acs_clean.rds\n",
      "                        acs_clean.csv\n",
      sep = "")
} else {
  cat("Output files already exist - skipping export.\n")
}
```

\newpage

# Join and Aggregate Data

## Convert LAPD to Spatial Points

```{r convert-and-align}
##convert lapd lat/lon to sf point geometry using WGS84 (EPSG:4326)
lapd_sf <- lapd |>
  st_as_sf(coords = c("lon_clean", "lat_clean"), crs = 4326)

##align crs if needed
if (st_crs(lapd_sf) != st_crs(acs)) {
  acs <- st_transform(acs, st_crs(lapd_sf))
}
```

## Spatial Join

```{r spatial-join}
##join each lapd incident to the census tract it falls within
lapd_joined <- st_join(lapd_sf, acs |> select(GEOID, NAME), left = TRUE)

##check how many incidents didn't match a tract
n_unmatched <- sum(is.na(lapd_joined$GEOID))
cat("Incidents matched to a tract:   ", nrow(lapd_joined) - n_unmatched, "\n",
    "Incidents with no tract match:  ", n_unmatched, "\n", 
    sep = "")
```

## Build RQ1 Incident-Level Dataset

RQ1 — *How do neighborhood-level socioeconomic characteristics predict whether a crime incident is violent or property-related?* — requires an incident-level dataset where each row is a single crime, the outcome is a binary violent/property flag, and the ACS socioeconomic characteristics of the incident's census tract serve as predictors. Drug, weapons-only, and other incidents that are neither violent nor property crimes are excluded.

```{r rq1-build}
##join acs socioeconomic predictors to each incident by tract geoid
##exclude incidents classified as Other (not violent or property)
incident_data <- lapd_joined |>
  st_drop_geometry() |>
  filter(!is.na(GEOID), crime_type != "Other") |>
  left_join(
    acs |>
      st_drop_geometry() |>
      select(
        GEOID,
        ##economic vulnerability
        median_income, gini, poverty_rate, unemployment_rate,
        no_vehicle_rate, housing_burden_rate,
        ##neighborhood stability
        vacancy_rate, renter_rate,
        ##social disorganization
        single_parent_rate, lang_isolation_rate, internet_rate,
        ##demographics
        total_pop, median_age,
        pct_white, pct_black, pct_asian, pct_hispanic,
        ##education and disability
        edu_rate_no_hsdip, edu_rate_college,
        disability_rate
      ),
    by = "GEOID"
  ) |>
  ##drop rows where acs predictors couldn't be joined (tracts outside la county acs pull)
  filter(!is.na(median_income)) |>
  mutate(
    ##binary outcome: 1 = violent, 0 = property
    outcome = as.integer(is_violent),

    ##multiclass outcome: collapsed crime category as factor
    outcome_multi = factor(case_when(
      crime_category %in% c("Homicide", "Weapons", "Arson")   ~ "Other Violent",
      crime_category %in% c("Sexual Assault", "Sex Offenses") ~ "Sexual Offenses",
      TRUE                                                     ~ crime_category
    ))
  ) |>
  select(
    rpt_no, date_occ, crime_cd, crime_category,
    crime_type, is_violent, is_property, outcome, outcome_multi,
    GEOID,
    median_income, gini, poverty_rate, unemployment_rate,
    no_vehicle_rate, housing_burden_rate,
    vacancy_rate, renter_rate,
    single_parent_rate, lang_isolation_rate, internet_rate,
    total_pop, median_age,
    pct_white, pct_black, pct_asian, pct_hispanic,
    edu_rate_no_hsdip, edu_rate_college,
    disability_rate
  )

cat("RQ1 incident-level dataset:\n",
    "  Total incidents (violent + property): ", nrow(incident_data),      "\n",
    "  Violent:                              ", sum(incident_data$is_violent),  "\n",
    "  Property:                             ", sum(incident_data$is_property), "\n",
    "  Unique tracts:                        ", n_distinct(incident_data$GEOID), "\n",
    sep = "")
```

```{r rq1-export}
if (!file.exists("output/incident_data.rds")) {
  saveRDS(incident_data, "output/incident_data.rds")
  write_csv(incident_data, "output/incident_data.csv")
  cat("Files saved to /output:\n",
      "                         incident_data.rds\n",
      "                         incident_data.csv\n",
      sep = "")
} else {
  cat("Output files already exist - skipping export.\n")
}
```

\newpage

## Aggregate to Tract Level for RQ2

```{r join-aggregate}
crime_by_tract <- lapd_joined |>
  st_drop_geometry() |>
  filter(!is.na(GEOID)) |>
  group_by(GEOID) |>
  summarise(
    total_incidents      = n(),

    ##crime category counts
    n_homicide           = sum(crime_category == "Homicide",        na.rm = TRUE),
    n_robbery            = sum(crime_category == "Robbery",         na.rm = TRUE),
    n_assault            = sum(crime_category == "Assault",         na.rm = TRUE),
    n_simple_assault     = sum(crime_category == "Simple Assault",  na.rm = TRUE),
    n_sexual_assault     = sum(crime_category == "Sexual Assault",  na.rm = TRUE),
    n_sex_offenses       = sum(crime_category == "Sex Offenses",    na.rm = TRUE),
    n_burglary           = sum(crime_category == "Burglary",        na.rm = TRUE),
    n_theft              = sum(crime_category == "Theft",           na.rm = TRUE),
    n_vehicle_theft      = sum(crime_category == "Vehicle Theft",   na.rm = TRUE),
    n_shoplifting        = sum(crime_category == "Shoplifting",     na.rm = TRUE),
    n_vandalism          = sum(crime_category == "Vandalism",       na.rm = TRUE),
    n_drug               = sum(crime_category == "Drug/Narcotic",   na.rm = TRUE),
    n_weapons            = sum(crime_category == "Weapons",         na.rm = TRUE),
    n_arson              = sum(crime_category == "Arson",           na.rm = TRUE),
    n_other              = sum(crime_category == "Other",           na.rm = TRUE),

    ##violent crime count
    n_violent            = sum(is_violent, na.rm = TRUE),
   
    ##clearance counts
    n_arrested      = sum(case_status %in% c("Adult Arrest", "Juv Arrest"), na.rm = TRUE),
    n_cleared_other = sum(case_status %in% c("Adult Other",  "Juv Other"),  na.rm = TRUE),

    .groups = "drop"
  )
```

\newpage

## Compute Crime Rates

```{r join-rates}
crime_by_tract <- crime_by_tract |>
  left_join(acs |> st_drop_geometry() |> select(GEOID, total_pop), by = "GEOID") |>
  mutate(
    rate_total          = (total_incidents  / total_pop / 5) * 1000,
    rate_violent        = (n_violent        / total_pop / 5) * 1000,
    rate_homicide       = (n_homicide       / total_pop / 5) * 1000,
    rate_robbery        = (n_robbery        / total_pop / 5) * 1000,
    rate_assault        = (n_assault        / total_pop / 5) * 1000,
    rate_simple_assault = (n_simple_assault / total_pop / 5) * 1000,
    rate_sexual_assault = (n_sexual_assault / total_pop / 5) * 1000,
    rate_sex_offenses   = (n_sex_offenses   / total_pop / 5) * 1000,
    rate_burglary       = (n_burglary       / total_pop / 5) * 1000,
    rate_theft          = (n_theft          / total_pop / 5) * 1000,
    rate_vehicle_theft  = (n_vehicle_theft  / total_pop / 5) * 1000,
    rate_shoplifting    = (n_shoplifting    / total_pop / 5) * 1000,
    rate_vandalism      = (n_vandalism      / total_pop / 5) * 1000, 
    rate_drug           = (n_drug           / total_pop / 5) * 1000,
    rate_weapons        = (n_weapons        / total_pop / 5) * 1000,
    rate_arson          = (n_arson          / total_pop / 5) * 1000
  ) |>
  select(-total_pop)
```

## Join ACS Variables

```{r join-acs}
##join aggregated crime data to full acs dataset by tract
tract_data <- acs |>
  left_join(crime_by_tract, by = "GEOID") |>
  ##replace NA crime counts and rates for tracts with no matched incidents
  mutate(
    total_incidents = replace_na(total_incidents, 0),
    across(starts_with("n_"),    \(x) replace_na(x, 0)),
    across(starts_with("rate_"), \(x) replace_na(x, 0))
  )
```

## Filter to City of Los Angeles

```{r filter-city, message=FALSE, warning=FALSE}
##city of los angeles boundary
if (!file.exists("output/la_city_boundary.rds")) {
  la_city <- places(state = "CA", cb = TRUE) |>
    filter(NAME == "Los Angeles") |>
    st_transform(st_crs(tract_data))
  saveRDS(la_city, "output/la_city_boundary.rds")
  cat("City boundary downloaded and saved.\n")
} else {
  la_city <- readRDS("output/la_city_boundary.rds")
  cat("City boundary loaded from cache.\n")
}
```

\newpage

```{r filter-city-2}
##keep tracts that intersect the city boundary
n_before <- nrow(tract_data)
tract_data <- tract_data |>
  filter(lengths(st_intersects(geometry, la_city)) > 0)

cat("Tracts outside City of LA removed: ", n_before - nrow(tract_data), "\n",
    "Remaining tracts:                  ", nrow(tract_data), "\n",
     sep = "")
```

## Tract Flags

```{r flag-unusual-tracts}
##tracts where rates may be unstable
tract_data <- tract_data |>
  mutate(
    flag_small_pop    = total_pop < 500,
    flag_zero_crime   = total_incidents == 0,
    flag_extreme_rate = rate_total > mean(rate_total, na.rm = TRUE) + 
                        3 * sd(rate_total, na.rm = TRUE)
  )

cat("Small population tracts (<500):  ", sum(tract_data$flag_small_pop),   "\n",
    "Zero incident tracts:            ", sum(tract_data$flag_zero_crime),   "\n",
    "Extreme rate tracts (3 SD):      ", sum(tract_data$flag_extreme_rate), "\n",
    sep = "")
```

## Check Join Results

```{r join-check}
cat("Total tracts in final dataset:        ", nrow(tract_data), "\n",
    "Tracts with zero incidents:           ", sum(tract_data$total_incidents == 0,
                                              na.rm = TRUE), "\n",
    "Tracts with missing crime rates:      ", sum(is.na(tract_data$rate_total)),  "\n",
    sep = "")
```

\newpage

## Zero-Incident Tract Details

```{r tract-pop}
##inspect zero-incident tracts to assess cause
zero_crime_tracts <- tract_data |>
  st_drop_geometry() |>
  filter(flag_zero_crime) |>
  select(GEOID, NAME, total_pop, flag_small_pop) |>
  arrange(total_pop)

cat("Zero-incident tracts with small population (<500): ",
    sum(zero_crime_tracts$flag_small_pop), " of ",
    nrow(zero_crime_tracts), "\n",
    sep = "")

knitr::kable(zero_crime_tracts,
             col.names = c("GEOID", "Name", "Total Population", "Small Pop Flag"))

```
 
## Remove Zero-Incident and Small Population Tracts

```{r remove-zero-crime}
n_before <- nrow(tract_data)

tract_data <- tract_data |>
  filter(!flag_zero_crime, !flag_small_pop)

cat("Tracts removed: ", n_before - nrow(tract_data), "\n",
    "Remaining tracts:             ", nrow(tract_data), "\n",
    sep = "")
```

\newpage

# Variable Selection

## Crime Variables

```{r crime-var-selection}
##create aggregate property rate for rq2 spatial analysis
tract_data <- tract_data |>
  mutate(
    rate_property = rate_burglary + rate_theft + rate_vehicle_theft +
                    rate_shoplifting + rate_vandalism
  )

##drop individual category counts and rates not used in modeling or eda
crime_vars_drop <- c(
  "n_homicide", "n_robbery", "n_assault", "n_simple_assault",
  "n_sexual_assault", "n_sex_offenses", "n_burglary", "n_theft",
  "n_vehicle_theft", "n_shoplifting", "n_vandalism", "n_drug",
  "n_weapons", "n_arson", "n_other",
  
  "rate_homicide", "rate_robbery", "rate_assault", "rate_simple_assault",
  "rate_sexual_assault", "rate_sex_offenses", "rate_burglary", "rate_theft",
  "rate_vehicle_theft", "rate_shoplifting", "rate_vandalism",
  "rate_weapons", "rate_arson"
)

tract_data <- tract_data |>
  select(-any_of(crime_vars_drop))

cat("Variables dropped:          ", length(crime_vars_drop), "\n",
    "Remaining crime variables:  ", tract_data |> st_drop_geometry() |>
      select(starts_with("n_"), starts_with("rate_"), "total_incidents") |>
      ncol(), "\n",
    sep = "")
```

## ACS Variables

### Check Near Zero Variance

```{r acs-near-zero-variance}
##check variance of all acs predictor rates across tracts
acs_predictors <- tract_data |>
  st_drop_geometry() |>
  select(
    median_income, gini, poverty_rate, unemployment_rate,
    no_vehicle_rate, housing_burden_rate, vacancy_rate, renter_rate,
    building_units, single_parent_rate, lang_isolation_rate, internet_rate,
    total_pop, median_age, household_size,
    pct_white, pct_black, pct_native, pct_asian, pct_pacific,
    pct_other, pct_two_or_more, pct_hispanic,
    edu_rate_no_hsdip, edu_rate_hs, edu_rate_some_col, edu_rate_college,
    disability_rate
  )

##compute variance and basic distribution stats for each variable
var_summary <- acs_predictors |>
  summarise(across(everything(), list(
    variance  = \(x) var(x, na.rm = TRUE),
    sd        = \(x) sd(x, na.rm = TRUE),
    mean      = \(x) mean(x, na.rm = TRUE),
    pct_zero  = \(x) mean(x < 0.01, na.rm = TRUE)
  ))) |>
  pivot_longer(
    everything(),
    names_to  = c("variable", ".value"),
    names_pattern = "(.*)_(variance|sd|mean|pct_zero)"
  ) |>
  arrange(variance)

knitr::kable(var_summary, 
             digits = 4,
             col.names = c("Variable", "Variance", "SD", "Mean", "Proportion Near Zero"))
```

\newpage

### Check Highly Correlated Variables

```{r acs-correlation}
acs_pred <- tract_data |>
  st_drop_geometry() |>
  select(
    median_income, gini, poverty_rate, unemployment_rate,
    no_vehicle_rate, housing_burden_rate, vacancy_rate, renter_rate,
    single_parent_rate, lang_isolation_rate, internet_rate,
    median_age, household_size,
    pct_white, pct_black, pct_two_or_more, pct_asian, pct_hispanic,
    edu_rate_no_hsdip, edu_rate_hs, edu_rate_some_col, edu_rate_college,
    disability_rate
  )

acs_cor <- cor(acs_pred, use = "pairwise.complete.obs")

##flag highly correlated pairs
acs_cor_long <- acs_cor |>
  as.data.frame() |>
  rownames_to_column("var1") |>
  pivot_longer(-var1, names_to = "var2", values_to = "correlation") |>
  filter(var1 < var2, abs(correlation) > 0.70) |>
  arrange(desc(abs(correlation)))

cat("Highly correlated pairs (|r| > 0.70):\n")
knitr::kable(acs_cor_long,
             digits = 3,
             col.names = c("Variable 1", "Variable 2", "Correlation"))
```

\newpage

### Select ACS Variables 

```{r acs-var-selection}
acs_vars_drop <- c(
  ##near-zero variance
  "building_units",
  "pct_pacific",
  "pct_native",
  "pct_other",
  "pct_two_or_more",
  
  ##multicollinearity / redundancy
  "household_size",
  "edu_rate_hs",
  "edu_rate_some_col"
)

tract_data <- tract_data |>
  select(-any_of(acs_vars_drop))

cat("ACS variables dropped:     ", length(acs_vars_drop), "\n",
    "Remaining ACS predictors:  ", tract_data |>
      st_drop_geometry() |>
      select(median_income, gini, poverty_rate, unemployment_rate,
             no_vehicle_rate, housing_burden_rate, vacancy_rate, renter_rate,
             single_parent_rate, lang_isolation_rate, internet_rate,
             total_pop, median_age, pct_white, pct_black, pct_asian, 
             pct_hispanic, edu_rate_no_hsdip, edu_rate_college,
             disability_rate) |>
      ncol(), "\n",
    sep = "")
```

## Export Dataset

```{r join-export, echo = FALSE}
if (!file.exists("output/tract_data.rds")) {
  saveRDS(tract_data, "output/tract_data.rds")
  tract_data |>
    st_drop_geometry() |>
    write_csv("output/tract_data.csv")
  cat("Files saved to /output:\n",
      "                        tract_data.rds\n",
      "                        tract_data.csv\n",
      sep = "")
} else {
  cat("Output files already exist - skipping export.\n")
}
```

\newpage

# Final Dataset Overview

```{r overview-dims}
cat("incident_data (incident-level, RQ1):\n",
    "  Rows:    ", nrow(incident_data),    "\n",
    "  Columns: ", ncol(incident_data),    "\n\n",
    "tract_data (tract-level, RQ2):\n",
    "  Rows:    ", nrow(tract_data),  "\n",
    "  Columns: ", ncol(tract_data),  "\n",
    sep = "")
```

## Tract Flags

```{r overview-flags, echo=FALSE}
tibble(
  Variable = c(
    "flag_small_pop",
    "flag_zero_crime",
    "flag_extreme_rate"
  ),
  Description = c(
    "Tract population below 500 (crime rates may be unstable)",
    "Tract had zero matched incidents 2020–2024",
    "Total crime rate exceeds 3 standard deviations above the mean"
  )
) |>
  knitr::kable(col.names = c("Variable", "Description"))
```

**Note:** 58 of 59 zero-crime tracts have populations above 500, suggesting 
these zeros reflect a spatial join artifact at the city boundary rather than 
genuine absence of crime. These tracts were excluded from the final dataset. Tracts with fewer than 500 residents were also excluded due to unstable crime rates.

## Tract-Level Crime Variables (`tract_data`)

```{r overview-crime, echo=FALSE}
tibble(
  Variable = c(
    ##spatial analysis outcomes
    "total_incidents", "n_violent",
    ##rates used in spatial analysis
    "rate_total", "rate_violent", "rate_property", "rate_drug",
    ##descriptive
    "n_arrested", "n_cleared_other"
  ),
  Type = c(
    rep("Count", 2),
    rep("Rate (per 1,000)", 4),
    rep("Count", 2)
  ),
  Description = c(
    "Total incidents 2020–2024",
    "Total violent crime incidents",
    "Total incident rate",
    "Violent incident rate",
    "Property crime rate",
    "Drug/narcotic incident rate",
    "Total arrests",
    "Total cases closed for other reasons"
  )
) |>
  knitr::kable(col.names = c("Variable", "Type", "Description")) |>
  kableExtra::pack_rows("Spatial Analysis Outcomes", 1, 2) |>
  kableExtra::pack_rows("Spatial Analysis Rates", 3, 6) |>
  kableExtra::pack_rows("Descriptive", 7, 8)
```

\newpage

## RQ1 Incident-Level Variables (`incident_data`)

```{r overview-rq1, echo=FALSE}
tibble(
  Variable = c(
    "rpt_no", "date_occ",
    "crime_cd", "crime_category", "crime_type",
    "is_violent", "is_property", "outcome", "outcome_multi",
    "GEOID",
    "median_income", "gini", "poverty_rate", "unemployment_rate",
    "no_vehicle_rate", "housing_burden_rate",
    "vacancy_rate", "renter_rate",
    "single_parent_rate", "lang_isolation_rate", "internet_rate",
    "total_pop", "median_age",
    "pct_white", "pct_black", "pct_asian", "pct_hispanic",
    "edu_rate_no_hsdip", "edu_rate_college",
    "disability_rate"
  ),
  Description = c(
    "LAPD report number",
    "Date of occurrence",
    "LAPD crime code",
    "Detailed crime category",
    "Violent / Property",
    "TRUE if violent crime",
    "TRUE if property crime",
    "Binary outcome: 1 = violent, 0 = property",
    "Multiclass outcome: specific crime category as factor",
    "Census tract GEOID (links to ACS predictors)",
    "Median household income",
    "Gini coefficient",
    "Proportion population below poverty line",
    "Proportion labor force unemployed",
    "Proportion households with no vehicle",
    "Proportion 50%+ of income on housing",
    "Proportion housing units vacant",
    "Proportion housing units renter-occupied",
    "Proportion households with single parent",
    "Proportion with limited English proficiency",
    "Proportion households with internet access",
    "Total tract population",
    "Median age of residents",
    "Proportion non-Hispanic white",
    "Proportion non-Hispanic Black",
    "Proportion non-Hispanic Asian",
    "Proportion Hispanic or Latino",
    "Proportion with less than high school diploma",
    "Proportion with bachelor's degree or higher",
    "Proportion with a disability"
  )
) |>
  knitr::kable(col.names = c("Variable", "Description")) |>
  kableExtra::pack_rows("Incident Identifiers", 1, 2) |>
  kableExtra::pack_rows("Crime Classification", 3, 5) |>
  kableExtra::pack_rows("Outcome Variables", 6, 9) |>
  kableExtra::pack_rows("Geography", 10, 10) |>
  kableExtra::pack_rows("Economic Vulnerability", 11, 16) |>
  kableExtra::pack_rows("Neighborhood Stability", 17, 18) |>
  kableExtra::pack_rows("Social Disorganization", 19, 21) |>
  kableExtra::pack_rows("Demographics", 22, 27) |>
  kableExtra::pack_rows("Education & Disability", 28, 30)
```

\newpage

## Tract-Level ACS Variables (`tract_data`)

```{r overview-acs, echo=FALSE}
tibble(
  Variable = c(
    ##economic vulnerability
    "median_income", "gini", "poverty_rate", "unemployment_rate",
    "no_vehicle_rate", "housing_burden_rate",
    ##neighborhood stability
    "vacancy_rate", "renter_rate",
    ##social disorganization
    "single_parent_rate", "lang_isolation_rate", "internet_rate",
    ##demographics
    "total_pop", "median_age",
    "pct_white", "pct_black", "pct_asian", "pct_hispanic",
    ##education
    "edu_rate_no_hsdip", "edu_rate_college",
    ##disability
    "disability_rate"
  ),
  Description = c(
    "Median household income",
    "Gini coefficient (0 = equal, 1 = unequal)",
    "Proportion population below poverty line",
    "Proportion labor force unemployed",
    "Proportion households with no vehicle",
    "Proportion 50%+ of income on housing",
    "Proportion housing units vacant",
    "Proportion housing units renter-occupied",
    "Proportion households with single parent",
    "Proportion with limited English proficiency",
    "Proportion households with internet access",
    "Total tract population",
    "Median age of residents",
    "Proportion non-Hispanic white",
    "Proportion non-Hispanic Black",
    "Proportion non-Hispanic Asian",
    "Proportion Hispanic or Latino",
    "Proportion with less than high school diploma",
    "Proportion with bachelor's degree or higher",
    "Proportion with a disability"
  )
) |>
  knitr::kable(col.names = c("Variable", "Description")) |>
  kableExtra::pack_rows("Economic Vulnerability", 1, 6) |>
  kableExtra::pack_rows("Neighborhood Stability", 7, 8) |>
  kableExtra::pack_rows("Social Disorganization", 9, 11) |>
  kableExtra::pack_rows("Demographics", 12, 17) |>
  kableExtra::pack_rows("Education", 18, 19) |>
  kableExtra::pack_rows("Disability", 20, 20)
```
