---
title: "EDA"
output:
  pdf_document:
    toc: true
    toc_depth: 3
    number_sections: true
    latex_engine: xelatex
date: "`r Sys.Date()`"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
library(tidyverse)
library(knitr)
library(kableExtra)
library(sf)
library(here)
library(tmap)

tmap_mode("plot")

##IBM Carbon colorblind-safe palette
ibm_blue    <- "#648FFF"
ibm_purple  <- "#785EF0"
ibm_magenta <- "#DC267F"
ibm_orange  <- "#FE6100"
ibm_gold    <- "#FFB000"

##load incident-level data for rq1
incident_data <- readRDS(here("output", "incident_data.rds"))

##load tract data for rq2
tract_data <- readRDS(here("output", "tract_data.rds"))
tract_data <- tract_data |>
  mutate(arrest_ratio = n_arrested / total_incidents)

##crime labels
crime_rate_labels <- c(
  rate_total      = "Total Crime",
  rate_violent    = "Violent Crime",
  rate_property   = "Property Crime",
  rate_drug       = "Drug Crime",
  total_incidents = "Total Incidents",
  n_violent       = "Violent Incidents",
  n_arrested      = "Arrests",
  n_cleared_other = "Cleared (Other)"
)

##acs labels
acs_plot_labels <- c(
  median_income       = "Median Income",
  gini                = "Gini Coefficient",
  poverty_rate        = "Poverty Rate",
  unemployment_rate   = "Unemployment Rate",
  no_vehicle_rate     = "No Vehicle Rate",
  housing_burden_rate = "Housing Burden Rate",
  vacancy_rate        = "Vacancy Rate",
  renter_rate         = "Renter Rate",
  single_parent_rate  = "Single-Parent Rate",
  lang_isolation_rate = "Language Isolation Rate",
  internet_rate       = "Internet Access Rate",
  disability_rate     = "Disability Rate",
  total_pop           = "Total Population",
  median_age          = "Median Age",
  pct_white           = "% White",
  pct_black           = "% Black",
  pct_asian           = "% Asian",
  pct_hispanic        = "% Hispanic",
  edu_rate_no_hsdip   = "No HS Diploma Rate",
  edu_rate_college    = "College Degree Rate"
)

##crime rate vars vector
crime_rate_vars <- c("rate_total", "rate_violent", "rate_property", "rate_drug")

##acs vars vector
acs_hist_vars <- c(
  "median_income", "gini", "poverty_rate", "unemployment_rate",
  "no_vehicle_rate", "housing_burden_rate", "vacancy_rate", "renter_rate",
  "single_parent_rate", "lang_isolation_rate", "internet_rate", "disability_rate",
  "pct_white", "pct_black", "pct_asian", "pct_hispanic",
  "median_age", "edu_rate_no_hsdip", "edu_rate_college"
)

kable_style <- function(kbl) {
  kbl |>
    kableExtra::kable_styling(
      latex_options = c("striped", "hold_position", "scale_down"),
      stripe_color  = "#f5f5f5",
      font_size     = 10
    )
}

##build Subtype color palette in descending size order
property_order <- incident_data |>
  filter(crime_type == "Property") |>
  count(outcome_multi) |>
  arrange(desc(n)) |>
  pull(outcome_multi)

violent_order <- incident_data |>
  filter(crime_type == "Violent") |>
  count(outcome_multi) |>
  arrange(desc(n)) |>
  pull(outcome_multi)

subtype_colors <- setNames(
  c(ibm_purple, ibm_blue,
    colorRampPalette(c(ibm_blue, "#00B4D8"))(3)[2],
    "#00B4D8", "#00CEC8"),
  property_order
) |>
  c(setNames(
    colorRampPalette(c(ibm_magenta, ibm_orange, ibm_gold))(4),
    violent_order
  ))

##ggplot theme
theme_ibm <- function() {
  theme_light() +
  theme(
    plot.title       = element_text(face = "bold", size = 13),
    plot.subtitle    = element_text(size = 9, color = "gray40"),
    strip.background = element_rect(fill = "#E8EEFF", color = NA),
    strip.text       = element_text(face = "bold", size = 10, color = "#1a1a1a"),
    axis.text        = element_text(size = 8),
    panel.grid.minor = element_blank(),
    panel.grid.major = element_line(color = "gray90")
  )
}

##boxplot outlier styling
boxplot_outlier <- list(
  outlier.shape = 21, outlier.fill = "white",
  outlier.color = "gray30", outlier.size = 0.8,
  outlier.alpha = 0.5
)

##heatmap theme and fill scale
theme_heatmap <- theme(
  axis.text.x = element_text(angle = 45, hjust = 1, size = 7),
  axis.text.y = element_text(size = 7),
  panel.grid  = element_blank()
)

heatmap_fill <- scale_fill_gradient2(
  low      = ibm_purple,
  mid      = "white",
  high     = ibm_gold,
  midpoint = 0,
  limits   = c(-1, 1)
)

##choropleth bbox, layout, and styling
la_bbox <- st_bbox(c(xmin = -118.95, ymin = 33.70,
                     xmax = -118.10, ymax = 34.35),
                   crs = st_crs(4326))

tm_map_layout <- function(title) {
  tm_layout(
    main.title      = title,
    main.title.size = 1,
    legend.outside  = TRUE,
    frame           = FALSE
  )
}

tm_quantile_poly <- function(col, title) {
  tm_polygons(
    col        = col,
    palette    = "plasma",
    style      = "quantile",
    n          = 5,
    title      = title,
    border.col = NA
  )
}

tm_choro <- function(data, col, legend_title, map_title) {
  tm_shape(data, bbox = la_bbox) +
    tm_quantile_poly(col, legend_title) +
    tm_map_layout(map_title)
}
```

\newpage

# Data Overview

## LAPD Crime Data 2020-2024

### Descriptive Statistics

```{r crime-desc-stats}
crime_desc <- tract_data |>
  st_drop_geometry() |>
  select(all_of(names(crime_rate_labels))) |>
  summarise(across(everything(), list(
    mean   = \(x) mean(x, na.rm = TRUE),
    median = \(x) median(x, na.rm = TRUE),
    sd     = \(x) sd(x, na.rm = TRUE),
    min    = \(x) min(x, na.rm = TRUE),
    max    = \(x) max(x, na.rm = TRUE)
  ))) |>
  pivot_longer(
    everything(),
    names_to  = c("variable", ".value"),
    names_pattern = "(.*)_(mean|median|sd|min|max)"
  ) |>
  mutate(variable = crime_rate_labels[variable])

knitr::kable(crime_desc,
             digits    = 2,
             col.names = c("Variable", "Mean", "Median", "SD", "Min", "Max"),
             booktabs  = TRUE,
             linesep   = "") |>
  kable_style() |>
  kableExtra::pack_rows("Rates (per 1,000)", 1, 4,
                        bold = TRUE, italic = FALSE, background = "#e8e8e8") |>
  kableExtra::pack_rows("Counts",            5, 8,
                        bold = TRUE, italic = FALSE, background = "#e8e8e8") |>
  kableExtra::column_spec(1, width = "5cm") |>
  kableExtra::column_spec(2:6, width = "1.6cm")
```

### Histograms

```{r crime-hist}
tract_data |>
  st_drop_geometry() |>
  select(all_of(crime_rate_vars)) |>
  pivot_longer(everything(), names_to = "variable", values_to = "value") |>
  mutate(variable = factor(crime_rate_labels[variable],
                           levels = crime_rate_labels[crime_rate_vars])) |>
  ggplot(aes(x = value)) +
  geom_histogram(bins = 40, fill = ibm_blue, color = "white") +
  facet_wrap(~ variable, scales = "free") +
  labs(
    title    = "Distribution of Annual Crime Rates Across Census Tracts",
    subtitle = "All distributions are right-skewed, indicating a small number of high-crime tracts",
    x        = "Rate (incidents per 1,000 residents per year)",
    y        = "Number of Tracts"
  ) +
  theme_ibm()
```

### Crime Breakdown

```{r crime-donut}
donut_colors <- c(
  "Property" = ibm_purple,
  "Violent"  = ibm_magenta,
  subtype_colors
)

##inner ring: crime type totals — factored to control stacking order
inner <- incident_data |>
  count(crime_type) |>
  mutate(crime_type = factor(crime_type, levels = c("Property", "Violent"))) |>
  arrange(crime_type) |>
  mutate(
    fraction = n / sum(n),
    ymax     = cumsum(fraction),
    ymin     = lag(ymax, default = 0),
    label    = paste0(crime_type, "\n", formatC(n, format = "d", big.mark = ","))
  )

##outer ring: factored to match inner ring stacking order
outer <- incident_data |>
  count(crime_type, outcome_multi) |>
  mutate(
    fraction   = n / sum(n),
    crime_type = factor(crime_type, levels = c("Property", "Violent"))
  ) |>
  arrange(crime_type, desc(n)) |>
  mutate(
    ymax = cumsum(fraction),
    ymin = lag(ymax, default = 0)
  )

ggplot() +
  ##outer ring
  geom_rect(data = outer,
            aes(ymin = ymin, ymax = ymax, xmin = 3.05, xmax = 4.1,
                fill = outcome_multi)) +
  ##inner ring
  geom_rect(data = inner,
            aes(ymin = ymin, ymax = ymax, xmin = 1.4, xmax = 2.95,
                fill = crime_type)) +
  scale_fill_manual(
    values = donut_colors,
    breaks = c(property_order, violent_order)
  ) +
  coord_polar(theta = "y") +
  xlim(c(0, 4.1)) +
  geom_text(data = inner,
            aes(x = 2.2, y = (ymin + ymax) / 2, label = label),
            size = 3.5, fontface = "bold", color = "white") +
  labs(
    title   = "Incident Composition by Crime Type and Subtype",
    caption = "Other Violent includes Homicide, Weapons, and Arson. Sexual Offenses includes Sexual Assault and Sex Offenses.",
    fill    = "Subtype"
  ) +
  theme_void() +
  theme(
    plot.title   = element_text(face = "bold", size = 13, hjust = 0.5),
    plot.caption = element_text(size = 8, color = "gray40", hjust = 0.5,
                                margin = margin(t = 10)),
    plot.margin  = margin(10, 10, 10, 50),
    legend.title = element_text(face = "bold")
  )
```

\newpage

## 2024 5-year American Census Survey Data

### Descriptive Statistics

```{r acs-desc-stats}
##compute descriptive statistics for acs predictor variables
acs_desc <- tract_data |>
  st_drop_geometry() |>
  select(all_of(names(acs_plot_labels))) |>
  summarise(across(everything(), list(
    mean   = \(x) mean(x, na.rm = TRUE),
    median = \(x) median(x, na.rm = TRUE),
    sd     = \(x) sd(x, na.rm = TRUE),
    min    = \(x) min(x, na.rm = TRUE),
    max    = \(x) max(x, na.rm = TRUE)
  ))) |>
  pivot_longer(
    everything(),
    names_to  = c("variable", ".value"),
    names_pattern = "(.*)_(mean|median|sd|min|max)"
  ) |>
  mutate(
    variable = acs_plot_labels[variable],
    across(c(mean, median, sd, min, max), \(x) {
      ifelse(variable == "Median Income",
             formatC(x, format = "f", digits = 0, big.mark = ","),
             formatC(x, format = "f", digits = 3))
    })
  )

knitr::kable(acs_desc,
             col.names = c("Variable", "Mean", "Median", "SD", "Min", "Max"),
             booktabs  = TRUE,
             linesep   = "",
             align     = c("l", "r", "r", "r", "r", "r")) |>
  kable_style() |>
  kableExtra::pack_rows("Economic Vulnerability", 1,  6,
                        bold = TRUE, background = "#e8e8e8") |>
  kableExtra::pack_rows("Neighborhood Stability",  7,  8,
                        bold = TRUE, background = "#e8e8e8") |>
  kableExtra::pack_rows("Social Disorganization",  9, 12,
                        bold = TRUE, background = "#e8e8e8") |>
  kableExtra::pack_rows("Demographics",           13, 18,
                        bold = TRUE, background = "#e8e8e8") |>
  kableExtra::pack_rows("Education",              19, 20,
                        bold = TRUE, background = "#e8e8e8") |>
  kableExtra::column_spec(1, width = "5.5cm") |>
  kableExtra::column_spec(2:6, width = "1.6cm")
```

### Histograms

```{r acs-hist, fig.height=8.5}
##all acs variable distributions in one faceted plot
tract_data |>
  st_drop_geometry() |>
  select(all_of(acs_hist_vars)) |>
  pivot_longer(everything(), names_to = "variable", values_to = "value") |>
  mutate(variable = factor(acs_plot_labels[variable],
                           levels = acs_plot_labels[acs_hist_vars])) |>
  ggplot(aes(x = value)) +
  geom_histogram(bins = 40, fill = ibm_blue, color = "white") +
  facet_wrap(~ variable, scales = "free", ncol = 3) +
  labs(
    title = "Distribution of ACS Predictor Variables Across Census Tracts",
    x     = "Value",
    y     = "Number of Tracts"
  ) +
  theme_ibm() +
  theme(axis.text.x = element_text(size = 6),
        axis.text.y = element_text(size = 6))
```

## Correlation Heatmaps

### Multicollinearity

```{r acs-correlation-heatmap, fig.height=9, fig.width=9}
##spearman correlation doesn't assume a linear relationship
tract_data |>
  st_drop_geometry() |>
  select(all_of(acs_hist_vars)) |>
  cor(use = "pairwise.complete.obs", method = "spearman") |>
  as.data.frame() |>
  rownames_to_column("var1") |>
  pivot_longer(-var1, names_to = "var2", values_to = "correlation") |>
  mutate(across(c(var1, var2), ~ acs_plot_labels[.])) |>
  ggplot(aes(x = var1, y = var2, fill = correlation)) +
  geom_tile(color = "white") +
  geom_text(aes(label = round(correlation, 2)), size = 2.2) +
  heatmap_fill +
  labs(
    title = "Spearman Correlations Among ACS Predictor Variables",
    x     = NULL,
    y     = NULL,
    fill  = "Correlation"
  ) +
  theme_ibm() +
  theme_heatmap
```

### ACS Predictors vs. Crime Outcomes

```{r correlation-heatmap}
tract_data |>
  st_drop_geometry() |>
  select(all_of(acs_hist_vars), rate_violent, rate_property, rate_total) |>
  cor(use = "pairwise.complete.obs", method = "spearman") |>
  as.data.frame() |>
  rownames_to_column("var1") |>
  pivot_longer(-var1, names_to = "var2", values_to = "correlation") |>
  filter(
    var1 %in% acs_hist_vars,
    var2 %in% c("rate_violent", "rate_property")
  ) |>
  mutate(
    var1 = acs_plot_labels[var1],
    var2 = crime_rate_labels[var2]
  ) |>
  ggplot(aes(x = var2, y = var1, fill = correlation)) +
  geom_tile() +
  geom_text(aes(label = round(correlation, 2)), size = 3) +
  heatmap_fill +
  theme_heatmap +
  labs(
    title = "Correlation Between ACS Predictors and Crime Outcomes",
    x     = "Crime Outcome",
    y     = "ACS Predictor"
  )
```

# RQ1: How do neighborhood-level socioeconomic characteristics predict whether a crime incident is violent or property-related in the City of Los Angeles?

## Incident Descriptive Statistics

```{r rq1-desc-stats}
##property categories in size order
property_rows <- incident_data |>
  filter(crime_type == "Property") |>
  count(crime_category) |>
  arrange(desc(n)) |>
  mutate(pct = n / nrow(incident_data) * 100)

##property total
property_total <- tibble(
  crime_category = "Total Property",
  n              = sum(property_rows$n),
  pct            = sum(property_rows$pct)
)

##violent categories in size order
violent_rows <- incident_data |>
  filter(crime_type == "Violent") |>
  count(crime_category) |>
  arrange(desc(n)) |>
  mutate(pct = n / nrow(incident_data) * 100)

##violent total
violent_total <- tibble(
  crime_category = "Total Violent",
  n              = sum(violent_rows$n),
  pct            = sum(violent_rows$pct)
)

##combine in correct order: property categories + total, then violent categories + total
bind_rows(
  property_rows,
  property_total,
  violent_rows,
  violent_total
) |>
  mutate(
    n   = formatC(n,   format = "d", big.mark = ","),
    pct = formatC(pct, format = "f", digits   = 1)
  ) |>
  knitr::kable(
    col.names = c("Subtype", "Count", "%"),
    booktabs  = TRUE,
    linesep   = "",
    align     = c("l", "r", "r")
  ) |>
  kable_style() |>
  kableExtra::pack_rows("Property", 1, nrow(property_rows) + 1) |>
  kableExtra::pack_rows("Violent",  nrow(property_rows) + 2,
                        nrow(property_rows) + nrow(violent_rows) + 2)
```

## Histograms

```{r incident-hist}
##distribution of crime type proportions across tracts
##proportions computed relative to violent + property + drug only
tract_data |>
  st_drop_geometry() |>
  mutate(
    rate_sum     = rate_violent + rate_property + rate_drug,
    pct_violent  = rate_violent  / rate_sum,
    pct_property = rate_property / rate_sum,
    pct_drug     = rate_drug     / rate_sum
  ) |>
  select(pct_violent, pct_property, pct_drug) |>
  pivot_longer(everything(),
               names_to  = "type",
               values_to = "proportion") |>
  mutate(type = case_when(
    type == "pct_violent"  ~ "Violent",
    type == "pct_property" ~ "Property",
    type == "pct_drug"     ~ "Drug"
  )) |>
  ggplot(aes(x = proportion, fill = type)) +
  geom_histogram(bins = 40, color = "white", alpha = 0.7,
                 position = "identity") +
  scale_fill_manual(values = c(
    "Violent"  = ibm_magenta,
    "Property" = ibm_purple,
    "Drug"     = ibm_gold
  )) +
  scale_x_continuous(labels = scales::percent) +
  labs(
    title    = "Distribution of Crime Type Proportions Across Census Tracts",
    subtitle = "Drug bar truncated at y = 250; full height = 1,140 tracts",
    x        = "Proportion of Violent, Property, and Drug Incidents",
    y        = "Number of Tracts",
    fill     = "Crime Type"
  ) +
  coord_cartesian(ylim = c(0, 250)) +
  theme_ibm()
```

```{r incident-hist-multi, fig.height=7}
##distribution of each crime Subtype proportion across tracts
incident_data |>
  group_by(GEOID) |>
  mutate(n_total = n()) |>
  filter(n_total >= 10) |>
  group_by(GEOID, outcome_multi, n_total) |>
  summarise(n_cat = n(), .groups = "drop") |>
  mutate(proportion = n_cat / n_total,
         outcome_multi = factor(outcome_multi, 
                                levels = c(property_order, violent_order))) |>
  ggplot(aes(x = proportion, fill = outcome_multi)) +
  geom_histogram(bins = 30, color = "white") +
  scale_fill_manual(values = subtype_colors) +
  scale_x_continuous(labels = scales::percent) +
  facet_wrap(~ outcome_multi, scales = "free", ncol = 3) +
  labs(
    title    = "Distribution of Crime Subtype Proportions Across Census Tracts",
    subtitle = "Tracts with fewer than 10 incidents excluded",
    x        = "Proportion of Incidents",
    y        = "Number of Tracts"
  ) +
  theme_ibm() +
  theme(legend.position = "none")
```

## Box Plot Comparisons by Crime Type

### Economic Vulnerability

```{r econ-box, fig.height=8}
##economic vulnerability variables by crime type at incident level
incident_data |>
  select(crime_type, median_income, gini, poverty_rate,
         unemployment_rate, no_vehicle_rate, housing_burden_rate) |>
  pivot_longer(-crime_type, names_to = "variable", values_to = "value") |>
  mutate(variable = acs_plot_labels[variable]) |>
  ggplot(aes(x = crime_type, y = value, fill = crime_type)) +
  do.call(geom_boxplot, boxplot_outlier) +
  scale_fill_manual(values = c("Violent" = ibm_magenta, "Property" = ibm_purple)) +
  facet_wrap(~ variable, scales = "free_y", ncol = 2) +
  labs(
    title = "Economic Vulnerability by Crime Type",
    x     = NULL,
    y     = "Value"
  ) +
  theme_ibm() +
  theme(legend.position = "none")
```

### Neighborhood Stability

```{r stab-box, fig.height=2.3}
##neighborhood stability variables by crime type at incident level
incident_data |>
  select(crime_type, vacancy_rate, renter_rate) |>
  pivot_longer(-crime_type, names_to = "variable", values_to = "value") |>
  mutate(variable = acs_plot_labels[variable]) |>
  ggplot(aes(x = crime_type, y = value, fill = crime_type)) +
  do.call(geom_boxplot, boxplot_outlier) +
  scale_fill_manual(values = c("Violent" = ibm_magenta, "Property" = ibm_purple)) +
  facet_wrap(~ variable, scales = "free_y", ncol = 2) +
  labs(
    title = "Neighborhood Stability by Crime Type",
    x     = NULL,
    y     = "Value"
  ) +
  theme_ibm() +
  theme(legend.position = "none")
```

### Social Disorganization

```{r soc-box, fig.height=4.6}
##social disorganization variables by crime type at incident level
incident_data |>
  select(crime_type, single_parent_rate, lang_isolation_rate,
         internet_rate, disability_rate) |>
  pivot_longer(-crime_type, names_to = "variable", values_to = "value") |>
  mutate(variable = acs_plot_labels[variable]) |>
  ggplot(aes(x = crime_type, y = value, fill = crime_type)) +
  do.call(geom_boxplot, boxplot_outlier) +
  scale_fill_manual(values = c("Violent" = ibm_magenta, "Property" = ibm_purple)) +
  facet_wrap(~ variable, scales = "free_y", ncol = 2) +
  labs(
    title = "Social Disorganization by Crime Type",
    x     = NULL,
    y     = "Value"
  ) +
  theme_ibm() +
  theme(legend.position = "none")
```

### Demographics

```{r demo-box, fig.height=8}
##demographic variables by crime type at incident level
incident_data |>
  select(crime_type, median_age,
         pct_white, pct_black, pct_asian, pct_hispanic) |>
  pivot_longer(-crime_type, names_to = "variable", values_to = "value") |>
  mutate(variable = acs_plot_labels[variable]) |>
  ggplot(aes(x = crime_type, y = value, fill = crime_type)) +
  do.call(geom_boxplot, boxplot_outlier) +
  scale_fill_manual(values = c("Violent" = ibm_magenta, "Property" = ibm_purple)) +
  facet_wrap(~ variable, scales = "free_y", ncol = 2) +
  labs(
    title = "Demographics by Crime Type",
    x     = NULL,
    y     = "Value"
  ) +
  theme_ibm() +
  theme(legend.position = "none")
```

### Educational Attainment

```{r edu-box, fig.height=2.3}
##education variables by crime type at incident level
incident_data |>
  select(crime_type, edu_rate_no_hsdip, edu_rate_college) |>
  pivot_longer(-crime_type, names_to = "variable", values_to = "value") |>
  mutate(variable = acs_plot_labels[variable]) |>
  ggplot(aes(x = crime_type, y = value, fill = crime_type)) +
  do.call(geom_boxplot, boxplot_outlier) +
  scale_fill_manual(values = c("Violent" = ibm_magenta, "Property" = ibm_purple)) +
  facet_wrap(~ variable, scales = "free_y", ncol = 2) +
  labs(
    title = "Educational Attainment by Crime Type",
    x     = NULL,
    y     = "Value"
  ) +
  theme_ibm() +
  theme(legend.position = "none")
```

### Multi Box

```{r multi-box, fig.height=4.6}
##top predictors by crime Subtype — establishes acs signal for multiclass modeling
incident_data |>
  select(outcome_multi, median_income, no_vehicle_rate, edu_rate_college) |>
  pivot_longer(-outcome_multi, names_to = "variable", values_to = "value") |>
  mutate(
    variable     = acs_plot_labels[variable],
    outcome_multi = factor(outcome_multi, levels = c(property_order, violent_order))
  ) |>
  ggplot(aes(x = outcome_multi, y = value, fill = outcome_multi)) +
  do.call(geom_boxplot, boxplot_outlier) +
  scale_fill_manual(values = subtype_colors) +
  facet_wrap(~ variable, scales = "free_y", ncol = 1) +
  labs(
    title = "Top ACS Predictors by Crime Subtype",
    x     = NULL,
    y     = "Value"
  ) +
  theme_ibm() +
  theme(
    legend.position = "none",
    axis.text.x     = element_text(angle = 25, hjust = 1)
  )
```

## Scatter Plots

```{r rq1-tract-setup}
##tract-level violent and property proportions for scatter plots
tract_violent <- incident_data |>
  group_by(GEOID) |>
  summarise(
    n_total      = n(),
    pct_violent  = mean(outcome),
    pct_property = 1 - mean(outcome),
    across(all_of(acs_hist_vars), first)
  ) |>
  filter(n_total >= 10)
```

### Economic Vulnerability

```{r econ-scatter, fig.height=6}
##economic vulnerability predictors vs tract violent and property proportions
tract_violent |>
  select(pct_violent, pct_property, median_income, gini, poverty_rate,
         unemployment_rate, no_vehicle_rate, housing_burden_rate) |>
  pivot_longer(c(pct_violent, pct_property),
               names_to = "outcome", values_to = "proportion") |>
  pivot_longer(c(median_income, gini, poverty_rate,
                 unemployment_rate, no_vehicle_rate, housing_burden_rate),
               names_to = "variable", values_to = "value") |>
  mutate(
    variable = acs_plot_labels[variable],
    outcome  = case_when(
      outcome == "pct_violent"  ~ "Violent",
      outcome == "pct_property" ~ "Property"
    )
  ) |>
  ggplot(aes(x = value, y = proportion, color = outcome)) +
  geom_point(alpha = 0.15, size = 0.5) +
  geom_smooth(method = "loess", se = TRUE, linewidth = 0.8) +
  scale_color_manual(values = c("Violent" = ibm_magenta, "Property" = ibm_purple)) +
  scale_y_continuous(labels = scales::percent) +
  facet_wrap(~ variable, scales = "free_x", ncol = 2) +
  labs(
    title = "Economic Vulnerability vs. Crime Type Proportion",
    x     = "Predictor Value",
    y     = "Proportion of Incidents",
    color = "Crime Type"
  ) +
  theme_ibm()
```

### Neighborhood Stability

```{r stab-scatter, fig.height=2}
##neighborhood stability predictors vs tract violent and property proportions
tract_violent |>
  select(pct_violent, pct_property, vacancy_rate, renter_rate) |>
  pivot_longer(c(pct_violent, pct_property),
               names_to = "outcome", values_to = "proportion") |>
  pivot_longer(c(vacancy_rate, renter_rate),
               names_to = "variable", values_to = "value") |>
  mutate(
    variable = acs_plot_labels[variable],
    outcome  = case_when(
      outcome == "pct_violent"  ~ "Violent",
      outcome == "pct_property" ~ "Property"
    )
  ) |>
  ggplot(aes(x = value, y = proportion, color = outcome)) +
  geom_point(alpha = 0.15, size = 0.5) +
  geom_smooth(method = "loess", se = TRUE, linewidth = 0.8) +
  scale_color_manual(values = c("Violent" = ibm_magenta, "Property" = ibm_purple)) +
  scale_y_continuous(labels = scales::percent) +
  facet_wrap(~ variable, scales = "free_x", ncol = 2) +
  labs(
    title = "Neighborhood Stability vs. Crime Type Proportion",
    x     = "Predictor Value",
    y     = "Proportion of Incidents",
    color = "Crime Type"
  ) +
  theme_ibm()
```

### Social Disorganization

```{r soc-scatter, fig.height=4}
##social disorganization predictors vs tract violent and property proportions
tract_violent |>
  select(pct_violent, pct_property, single_parent_rate, lang_isolation_rate,
         internet_rate, disability_rate) |>
  pivot_longer(c(pct_violent, pct_property),
               names_to = "outcome", values_to = "proportion") |>
  pivot_longer(c(single_parent_rate, lang_isolation_rate,
                 internet_rate, disability_rate),
               names_to = "variable", values_to = "value") |>
  mutate(
    variable = acs_plot_labels[variable],
    outcome  = case_when(
      outcome == "pct_violent"  ~ "Violent",
      outcome == "pct_property" ~ "Property"
    )
  ) |>
  ggplot(aes(x = value, y = proportion, color = outcome)) +
  geom_point(alpha = 0.15, size = 0.5) +
  geom_smooth(method = "loess", se = TRUE, linewidth = 0.8) +
  scale_color_manual(values = c("Violent" = ibm_magenta, "Property" = ibm_purple)) +
  scale_y_continuous(labels = scales::percent) +
  facet_wrap(~ variable, scales = "free_x", ncol = 2) +
  labs(
    title = "Social Disorganization vs. Crime Type Proportion",
    x     = "Predictor Value",
    y     = "Proportion of Incidents",
    color = "Crime Type"
  ) +
  theme_ibm()
```

### Demographics

```{r demo-scatter, fig.height=6}
##demographic predictors vs tract violent and property proportions
tract_violent |>
  select(pct_violent, pct_property, median_age,
         pct_white, pct_black, pct_asian, pct_hispanic) |>
  pivot_longer(c(pct_violent, pct_property),
               names_to = "outcome", values_to = "proportion") |>
  pivot_longer(c(median_age, pct_white,
                 pct_black, pct_asian, pct_hispanic),
               names_to = "variable", values_to = "value") |>
  mutate(
    variable = acs_plot_labels[variable],
    outcome  = case_when(
      outcome == "pct_violent"  ~ "Violent",
      outcome == "pct_property" ~ "Property"
    )
  ) |>
  ggplot(aes(x = value, y = proportion, color = outcome)) +
  geom_point(alpha = 0.15, size = 0.5) +
  geom_smooth(method = "loess", se = TRUE, linewidth = 0.8) +
  scale_color_manual(values = c("Violent" = ibm_magenta, "Property" = ibm_purple)) +
  scale_y_continuous(labels = scales::percent) +
  facet_wrap(~ variable, scales = "free_x", ncol = 2) +
  labs(
    title = "Demographics vs. Crime Type Proportion",
    x     = "Predictor Value",
    y     = "Proportion of Incidents",
    color = "Crime Type"
  ) +
  theme_ibm()
```

### Educational Attainment

```{r edu-scatter, fig.height=2}
##education predictors vs tract violent and property proportions
tract_violent |>
  select(pct_violent, pct_property, edu_rate_no_hsdip, edu_rate_college) |>
  pivot_longer(c(pct_violent, pct_property),
               names_to = "outcome", values_to = "proportion") |>
  pivot_longer(c(edu_rate_no_hsdip, edu_rate_college),
               names_to = "variable", values_to = "value") |>
  mutate(
    variable = acs_plot_labels[variable],
    outcome  = case_when(
      outcome == "pct_violent"  ~ "Violent",
      outcome == "pct_property" ~ "Property"
    )
  ) |>
  ggplot(aes(x = value, y = proportion, color = outcome)) +
  geom_point(alpha = 0.15, size = 0.5) +
  geom_smooth(method = "loess", se = TRUE, linewidth = 0.8) +
  scale_color_manual(values = c("Violent" = ibm_magenta, "Property" = ibm_purple)) +
  scale_y_continuous(labels = scales::percent) +
  facet_wrap(~ variable, scales = "free_x", ncol = 2) +
  labs(
    title = "Educational Attainment vs. Crime Type Proportion",
    x     = "Predictor Value",
    y     = "Proportion of Incidents",
    color = "Crime Type"
  ) +
  theme_ibm()
```

# RQ2: Are high-crime census tracts spatially clustered in the City of LA, and do neighboring tracts share similar socioeconomic characteristics?

## Arrest Ratio

### Descriptive Statistics

```{r arrest-ratio-stats}
tract_data |>
  st_drop_geometry() |>
  summarise(
    Mean   = mean(arrest_ratio, na.rm = TRUE),
    Median = median(arrest_ratio, na.rm = TRUE),
    SD     = sd(arrest_ratio, na.rm = TRUE),
    Min    = min(arrest_ratio, na.rm = TRUE),
    Max    = max(arrest_ratio, na.rm = TRUE)
  ) |>
  knitr::kable(
    digits   = 3,
    booktabs = TRUE,
    linesep  = "",
    align    = c("r", "r", "r", "r", "r")
  ) |>
  kable_style() |>
  kableExtra::add_header_above(
    c("Descriptive Statistics: Arrest-to-Incident Ratio" = 5),
    bold = TRUE, line = TRUE
  )
```

### Distribution

```{r arrest-ratio-dist}
tract_data |>
  st_drop_geometry() |>
  ggplot(aes(x = arrest_ratio)) +
  geom_histogram(bins = 40, fill = ibm_blue, color = "white") +
  scale_x_continuous(labels = scales::percent, limits = c(0, 0.4)) +
  labs(
    title    = "Distribution of Arrest-to-Incident Ratio Across Census Tracts",
    subtitle = "Most tracts cluster below 15%, with a small number of high-arrest-rate outliers",
    x        = "Arrest Ratio (Arrests / Total Incidents)",
    y        = "Number of Tracts"
  ) +
  theme_ibm()
```

### Scatter

```{r arrest-ratio-scatter, fig.height=8}
##arrest ratio vs socioeconomic predictors — policing equity context
tract_data |>
  st_drop_geometry() |>
  select(arrest_ratio, median_income, poverty_rate,
         no_vehicle_rate, pct_black, pct_hispanic, pct_white) |>
  pivot_longer(-arrest_ratio, names_to = "variable", values_to = "value") |>
  mutate(variable = acs_plot_labels[variable]) |>
  ggplot(aes(x = value, y = arrest_ratio)) +
  geom_point(alpha = 0.3, size = 0.8, color = ibm_blue) +
  geom_smooth(method = "loess", se = TRUE, color = ibm_magenta, linewidth = 0.8) +
  facet_wrap(~ variable, scales = "free_x", ncol = 2) +
  scale_y_continuous(labels = scales::percent) +
  labs(
    title = "Arrest-to-Incident Ratio vs. Socioeconomic Predictors",
    x     = "Predictor Value",
    y     = "Arrest Ratio (Arrests / Total Incidents)"
  ) +
  coord_cartesian(ylim = c(0, 0.4)) +
  theme_ibm()
```

## Crime Rate Maps

### Violent vs. Property Crime Across LA Census Tracts

```{r violent-prop-map}
##combined violent and property crime proportion maps
tract_violent_joined <- tract_data |>
  left_join(tract_violent |> select(GEOID, pct_violent, pct_property), by = "GEOID")

violent_map  <- tm_choro(tract_violent_joined, "pct_violent",  "Violent Proportion",  "Proportion Violent")
property_map <- tm_choro(tract_violent_joined, "pct_property", "Property Proportion", "Proportion Property")

tmap_arrange(violent_map, property_map, ncol = 2)
```

### Violent vs. Property Crime Across LA Census Tracts

```{r income-arrest-map}
##arrest ratio and median income maps side by side
arrest_map <- tm_choro(tract_data, "arrest_ratio", 
                       "Arrest Ratio", 
                       "Arrest-to-Incident Ratio Across LA Census Tracts")

income_map <- tm_choro(tract_data, "median_income",
                       "Median Income (USD)",
                       "Median Household Income Across LA Census Tracts")

tmap_arrange(arrest_map, income_map, ncol = 2)
```

