---
title: "RQ2"
output:
  pdf_document:
    toc: true
    toc_depth: 2
    number_sections: true
    latex_engine: xelatex
date: "`r Sys.Date()`"
---

\newpage

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE, fig.height = 5, fig.width = 8)
library(tidyverse)
library(knitr)
library(kableExtra)
library(sf)
library(here)
library(tmap)
library(spdep)
library(broom)
library(GWmodel)
library(sp)

tmap_mode("plot")

set.seed(490)

##IBM Carbon colorblind-safe palette
ibm_blue    <- "#648FFF"
ibm_purple  <- "#785EF0"
ibm_magenta <- "#DC267F"
ibm_orange  <- "#FE6100"
ibm_gold    <- "#FFB000"

crime_rate_labels <- c(
  log1p_rate_violent  = "Violent Crime",
  log1p_rate_property = "Property Crime"
)

acs_var_labels <- c(
  median_income       = "Median Income",
  gini                = "Gini Coefficient",
  poverty_rate        = "Poverty Rate",
  unemployment_rate   = "Unemployment Rate",
  no_vehicle_rate     = "No Vehicle Rate",
  housing_burden_rate = "Housing Burden Rate",
  vacancy_rate        = "Vacancy Rate",
  renter_rate         = "Renter Rate",
  single_parent_rate  = "Single-Parent Rate",
  lang_isolation_rate = "Language Isolation Rate",
  internet_rate       = "Internet Access Rate",
  disability_rate     = "Disability Rate",
  median_age          = "Median Age",
  pct_white           = "% White",
  pct_black           = "% Black",
  pct_asian           = "% Asian",
  pct_hispanic        = "% Hispanic",
  edu_rate_no_hsdip   = "No HS Diploma Rate",
  edu_rate_college    = "College Degree Rate"
)

##full acs predictor set
acs_model_vars <- c(
  "median_income", "gini", "poverty_rate", "unemployment_rate",
  "no_vehicle_rate", "housing_burden_rate", "vacancy_rate", "renter_rate",
  "single_parent_rate", "lang_isolation_rate", "internet_rate", "disability_rate",
  "median_age", "pct_white", "pct_black", "pct_asian", "pct_hispanic",
  "edu_rate_no_hsdip", "edu_rate_college"
)

##post-vif retained predictors
model_vars <- c(
  "median_income", "gini", "poverty_rate", "unemployment_rate",
  "no_vehicle_rate", "housing_burden_rate", "vacancy_rate", "renter_rate",
  "single_parent_rate", "lang_isolation_rate", "internet_rate", "disability_rate",
  "median_age", "edu_rate_no_hsdip"
)

##kable styling
kable_style <- function(kbl) {
  kbl |>
    kableExtra::kable_styling(
      latex_options = c("striped", "hold_position", "scale_down"),
      stripe_color  = "#f5f5f5",
      font_size     = 10
    )
}

##ggplot theme
theme_ibm <- function() {
  theme_light() +
  theme(
    plot.title       = element_text(face = "bold", size = 13),
    plot.subtitle    = element_text(size = 9, color = "gray40"),
    strip.background = element_rect(fill = "#E8EEFF", color = NA),
    strip.text       = element_text(face = "bold", size = 10, color = "#1a1a1a"),
    axis.text        = element_text(size = 8),
    panel.grid.minor = element_blank(),
    panel.grid.major = element_line(color = "gray90")
  )
}

##choropleths
la_bbox <- st_bbox(c(xmin = -118.95, ymin = 33.70,
                     xmax = -118.10, ymax = 34.35),
                   crs = st_crs(4326))

tm_map_layout <- function(title) {
  tm_layout(
    main.title      = title,
    main.title.size = 1,
    legend.outside  = TRUE,
    frame           = FALSE
  )
}

tm_choro <- function(col, legend_title, map_title) {
  tm_shape(tract_data, bbox = la_bbox) +
    tm_polygons(
      col        = col,
      palette    = "-PuOr",
      style      = "quantile",
      n          = 5,
      title      = legend_title,
      border.col = NA
    ) +
    tm_map_layout(map_title)
}

##aicc
ols_aicc <- function(model) {
  n <- nobs(model)
  k <- length(coef(model))
  AIC(model) + (2 * k^2 + 2 * k) / (n - k - 1)
}

##symmetric fixed breaks centered on zero (residuals, coefficients)
sym_breaks <- function(vals, n = 10) {
  lim <- max(abs(vals), na.rm = TRUE)
  seq(-lim, lim, length.out = n + 1)
}

##sequential fixed breaks from floor to max (local r2)
seq_breaks <- function(vals, n = 10) {
  lo <- min(0, min(vals, na.rm = TRUE))
  hi <- max(vals, na.rm = TRUE)
  seq(lo, hi, length.out = n + 1)
}

##lisa color palette 
lisa_palette <- c(
  "High-High"       = ibm_magenta,
  "High-Low"        = ibm_orange,
  "Low-High"        = ibm_blue,
  "Low-Low"         = ibm_purple,
  "Not Significant" = "#f0f0f0"
)

##lisa cluster map
lisa_map <- function(data, title, legend_title = "LISA Cluster") {
  data |>
    tm_shape(bbox = la_bbox) +
    tm_polygons(
      col          = "quad",
      palette      = lisa_palette,
      title        = legend_title,
      border.col   = NA,
      border.alpha = 0,
      colorNA      = "gray60",
      textNA       = "Missing",
      showNA       = TRUE
    ) +
    tm_map_layout(title)
}

#residuals map
resid_map    <- function(data, outcome_label, title, breaks = NULL)
  outcome_choro(data, "residual", "-PuOr", "Residual",  outcome_label, title, breaks = breaks)

##r2 map
local_r2_map <- function(data, outcome_label, title, breaks = NULL)
  outcome_choro(data, "local_r2", "viridis", "Local R²", outcome_label, title, breaks = breaks)

##placeholder panel
blank_panel <- tm_shape(
  st_sf(geometry = st_sfc(st_polygon(list(matrix(
    c(-118.95, 33.70, -118.10, 33.70, -118.10, 34.35,
      -118.95, 34.35, -118.95, 33.70),
    ncol = 2, byrow = TRUE))), crs = st_crs(4326))),
  bbox = la_bbox) +
  tm_fill(col = "white") +
  tm_layout(
    frame          = FALSE,
    frame.lwd      = 0,
    legend.show    = FALSE,
    bg.color       = "white",
    outer.bg.color = "white",
    outer.margins  = 0,
    inner.margins  = 0,
    asp            = 0
  )

##extract local coefficients
extract_coef_data <- function(sdf_data, outcome_label) {
  map_dfr(model_vars, function(var) {
    coef_val <- sdf_data[[var]]
    t_val    <- sdf_data[[paste0(var, "_TV")]]
    tract_data |>
      st_drop_geometry() |>
      select(GEOID) |>
      mutate(
        outcome   = outcome_label,
        predictor = acs_var_labels[var],
        coef      = if_else(abs(t_val) >= 1.96, coef_val, NA_real_)
      )
  })
}

##coefficient choropleth
coef_map <- function(map_data, var_label, outcome_label, title, breaks = NULL) {
  tract_data |>
    left_join(
      map_data |> filter(predictor == var_label, outcome == outcome_label),
      by = "GEOID"
    ) |>
    tm_shape(bbox = la_bbox) +
    tm_polygons(
      col          = "coef",
      palette      = "-PuOr",
      style        = if (is.null(breaks)) "quantile" else "fixed",
      breaks       = breaks,
      n            = if (is.null(breaks)) 5 else NULL,
      midpoint     = 0,
      title        = "Coefficient",
      border.col   = NA,
      border.alpha = 0,
      colorNA      = "gray80",
      textNA       = "Not Significant / Missing"
    ) +
    tm_map_layout(title)
}

##outcome choropleth 
outcome_choro <- function(data, col, palette, legend_title, outcome_label, title,
                          colorNA = "gray60", textNA = "Missing", breaks = NULL) {
  tract_data |>
    left_join(data |> filter(outcome == outcome_label), by = "GEOID") |>
    tm_shape(bbox = la_bbox) +
    tm_polygons(
      col          = col,
      palette      = palette,
      style        = if (is.null(breaks)) "quantile" else "fixed",
      breaks       = breaks,
      n            = if (is.null(breaks)) 5 else NULL,
      title        = legend_title,
      border.col   = NA,
      border.alpha = 0,
      colorNA      = colorNA,
      textNA       = textNA
    ) +
    tm_map_layout(title)
}

##mgwr coefficient summary
mgwr_coef_summary <- function(mgwr_model, outcome_label) {
  sdf <- mgwr_model$SDF@data
  map_dfr(model_vars, function(var) {
    coef_val <- sdf[[var]]
    t_val    <- sdf[[paste0(var, "_TV")]]
    sig      <- abs(t_val) >= 1.96
    tibble(
      Outcome         = outcome_label,
      Predictor       = acs_var_labels[var],
      Min             = round(min(coef_val,    na.rm = TRUE), 3),
      Mean            = round(mean(coef_val,   na.rm = TRUE), 3),
      Median          = round(median(coef_val, na.rm = TRUE), 3),
      Max             = round(max(coef_val,    na.rm = TRUE), 3),
      `% Significant` = paste0(round(mean(sig, na.rm = TRUE) * 100, 1), "%"),
      pct_sig_num     = mean(sig, na.rm = TRUE)
    )
  })
}

##local r2 for mgwr
mgwr_local_r2 <- function(mgwr_model, outcome_var, bw_k, dist_mat) {
  y     <- tract_data[[outcome_var]]
  resid <- mgwr_model$SDF$residual
  n     <- nrow(dist_mat)
  W     <- matrix(0, n, n)
  for (i in seq_len(n)) {
    d      <- dist_mat[i, ]
    h_i    <- sort(d)[bw_k + 1]
    w      <- ifelse(d < h_i, (1 - (d / h_i)^2)^2, 0)
    W[i, ] <- w
  }
  map_dbl(seq_len(n), function(i) {
    w       <- W[i, ]
    y_wmean <- sum(w * y) / sum(w)
    tss     <- sum(w * (y - y_wmean)^2)
    rss     <- sum(w * resid^2)
    if (tss < 1e-6) return(NA_real_)
    1 - rss / tss
  })
}
```

```{r load-data}
tract_data <- readRDS(here("output", "tract_data.rds"))
tract_data <- tract_data |>
  mutate(
    arrest_ratio        = n_arrested / total_incidents,
    ##log1p transformation applied to right-skewed crime rates
    log1p_rate_violent  = log1p(rate_violent),
    log1p_rate_property = log1p(rate_property)
    )
```

# Spatial Autocorrelation

## Spatial Weights

A weights matrix was constructed across all tracts using queen contiguity (any touching counts as a neighbor). One tract (GEOID 06037602201) has no contiguous neighbors, and is retained in modelling using `zero.policy = TRUE` and excluded from Bivariate LISA, which does not support that option.

```{r spatial-weights}
nb_queen <- poly2nb(tract_data, queen = TRUE)

##row-standardize weights so each tract's neighbors sum to 1
w_queen  <- nb2listw(nb_queen, style = "W", zero.policy = TRUE)

tibble(
  Metric = c("Number of Tracts", "Average Neighbors", "Isolated Tracts"),
  Value  = c("1,221", "6.15", "1")
) |>
  knitr::kable(
    caption = "Queen Contiguity Spatial Weights Matrix Summary",
    align   = c("l", "r")
  ) |>
  kable_style()

isolated_geoid <- tract_data$GEOID[card(nb_queen) == 0]
```

## Calculate Global Moran's I

```{r global-morans-i}
moran_results <- map_dfr(names(crime_rate_labels), function(var) {
  result <- moran.mc(tract_data[[var]], listw = w_queen,
                     nsim = 999, zero.policy = TRUE)
  tibble(
    Outcome     = crime_rate_labels[var],
    `Moran's I` = round(result$statistic, 4),
    `p-value`   = round(result$p.value,   4)
  )
})

moran_results |>
  knitr::kable(
    caption = "Global Moran's I: Crime Rate Spatial Autocorrelation",
    align   = c("l", "r", "r")
  ) |>
  kable_style()
```

## Visualize Global Moran's I

```{r global-morans-plot, fig.height=4, fig.width=7}
##standardize each rate then compute spatial lag from standardized values
moran_plot_data <- map_dfr(names(crime_rate_labels), function(var) {
  x     <- scale(tract_data[[var]])[, 1]
  lag_x <- lag.listw(w_queen, x, zero.policy = TRUE)
  tibble(
    x       = x,
    lag_x   = lag_x,
    outcome = crime_rate_labels[var]
  )
})

moran_plot_data |>
  ggplot(aes(x = x, y = lag_x)) +
  geom_hline(yintercept = 0, color = "gray40", linewidth = 0.5, linetype = "dashed") +
  geom_vline(xintercept = 0, color = "gray40", linewidth = 0.5, linetype = "dashed") +
  geom_point(alpha = 0.25, size = 0.8, color = ibm_blue) +
  geom_smooth(method = "lm", se = FALSE, color = ibm_magenta, linewidth = 0.8) +
  facet_wrap(~ outcome) +
  labs(
    title = "Moran Scatterplots: Spatial Autocorrelation of Crime Rates",
    x = "Standardized Crime Rate",
    y = "Spatially Lagged Crime Rate"
  ) +
  coord_cartesian(xlim = c(-2, 6), ylim = c(-2, 6)) +
  theme_ibm()
```

## Global Moran's I: Socioeconomic Predictors

```{r predictor-morans-i}
map_dfr(acs_model_vars, function(var) {
  result <- moran.mc(tract_data[[var]], listw = w_queen,
                     nsim = 999, zero.policy = TRUE)
  tibble(
    Predictor   = acs_var_labels[var],
    `Moran's I` = round(result$statistic, 4),
    `p-value`   = round(result$p.value,   4)
  )
}) |>
  arrange(desc(`Moran's I`)) |>
  knitr::kable(
    caption = "Global Moran's I: ACS Predictors",
    align   = c("l", "r", "r")
  ) |>
  kable_style()
```

\newpage

# Variable Inflation Factor (VIF)

All 19 candidate ACS predictors were screened for multicollinearity. Predictors with VIF > 10 were removed.

```{r vif}
fit <- lm(
  as.formula(paste("log1p_rate_violent ~", paste(acs_model_vars, collapse = " + "))),
  data = st_drop_geometry(tract_data) |>
    mutate(across(all_of(acs_model_vars), ~ as.numeric(scale(.x))))
)

vif_results <- tibble(
  Predictor = names(car::vif(fit)),
  VIF       = round(car::vif(fit), 3)
)

vif_results |>
  mutate(
    Predictor = acs_var_labels[Predictor],
    Decision  = if_else(VIF > 10, "Removed", "Retained")
  ) |>
  arrange(desc(VIF)) |>
  knitr::kable(
    caption   = "VIF Screening: Full Predictor Set",
    digits    = 3,
    col.names = c("Predictor", "VIF", "Decision"),
    align     = c("l", "r", "l")
  ) |>
  kable_style()
```

# Local Indicators of Spatial Association (LISA)

Global Moran's I confirmed that both crime rates and socioeconomic predictors cluster spatially, but does not identify where. LISA decomposes the global statistic to each tract. At 1,221 tests, the p < 0.05 threshold would produce approximately 61 expected false positives, compared to roughly 12 at p < 0.01.

```{r lisa}
lisa_results <- map(names(crime_rate_labels), function(var) {
  localmoran_perm(tract_data[[var]], listw = w_queen,
                  nsim = 999, zero.policy = TRUE)
}) |>
  set_names(names(crime_rate_labels))

lisa_classified <- map_dfr(names(crime_rate_labels), function(var) {
  x     <- scale(tract_data[[var]])[, 1]
  lag_x <- lag.listw(w_queen, x, zero.policy = TRUE)
  p_val <- lisa_results[[var]][, "Pr(z != E(Ii))"]

  tract_data |>
    st_drop_geometry() |>
    select(GEOID) |>
    mutate(
      outcome = crime_rate_labels[var],
      x       = x,
      lag_x   = lag_x,
      p_val   = p_val,
      quad    = case_when(
        p_val >= 0.01        ~ "Not Significant",
        x > 0 & lag_x > 0   ~ "High-High",
        x > 0 & lag_x <= 0  ~ "High-Low",
        x <= 0 & lag_x > 0  ~ "Low-High",
        x <= 0 & lag_x <= 0 ~ "Low-Low"
      )
    )
})
```

## Crime Outcomes

```{r lisa-table}
lisa_classified |>
  count(outcome, quad) |>
  pivot_wider(names_from = outcome, values_from = n, values_fill = 0) |>
  rename(Classification = quad) |>
  arrange(match(Classification,
                c("High-High", "High-Low", "Low-High", "Low-Low", "Not Significant"))) |>
  knitr::kable(
    caption = "LISA Classifications by Crime Rate Outcome",
    align   = c("l", "r", "r")
  ) |>
  kable_style()
```

\newpage

## Predictor LISA

```{r predictor-lisa-compute}
predictor_lisa_results <- map(model_vars, function(var) {
  localmoran_perm(tract_data[[var]], listw = w_queen,
                  nsim = 999, zero.policy = TRUE)
}) |>
  set_names(model_vars)

predictor_lisa_classified <- map_dfr(model_vars, function(var) {
  x     <- scale(tract_data[[var]])[, 1]
  lag_x <- lag.listw(w_queen, x, zero.policy = TRUE)
  p_val <- predictor_lisa_results[[var]][, "Pr(z != E(Ii))"]

  tract_data |>
    st_drop_geometry() |>
    select(GEOID) |>
    mutate(
      predictor = acs_var_labels[var],
      x         = x,
      lag_x     = lag_x,
      p_val     = p_val,
      quad      = case_when(
        p_val >= 0.01        ~ "Not Significant",
        x > 0 & lag_x > 0   ~ "High-High",
        x > 0 & lag_x <= 0  ~ "High-Low",
        x <= 0 & lag_x > 0  ~ "Low-High",
        x <= 0 & lag_x <= 0 ~ "Low-Low"
      )
    )
})

predictor_lisa_classified |>
  count(predictor, quad) |>
  pivot_wider(names_from = quad, values_from = n, values_fill = 0) |>
  rename(Predictor = predictor) |>
  select(Predictor, `High-High`, `High-Low`, `Low-High`, `Low-Low`, `Not Significant`) |>
  arrange(desc(`High-High`)) |>
  knitr::kable(
    caption = "LISA Classifications: ACS Predictors",
    align   = c("l", "r", "r", "r", "r", "r")
  ) |>
  kable_style()
```

## LISA Maps

### Crime vs Property

```{r lisa-map}
lisa_map_data <- tract_data |>
  left_join(lisa_classified, by = "GEOID") |>
  mutate(
    quad = factor(quad, levels = c("High-High", "High-Low",
                                   "Low-High", "Low-Low",
                                   "Not Significant"))
  )

tmap_arrange(
  lisa_map_data |> filter(outcome == "Violent Crime")  |> lisa_map("Violent Crime LISA Clusters"),
  lisa_map_data |> filter(outcome == "Property Crime") |> lisa_map("Property Crime LISA Clusters"),
  ncol = 2, outer.margins = 0
)
```

### Predictors

```{r predictor-lisa-maps, fig.height=4, fig.show='asis'}
predictor_lisa_map_data <- tract_data |>
  left_join(predictor_lisa_classified, by = "GEOID") |>
  mutate(
    quad = factor(quad, levels = c("High-High", "High-Low",
                                   "Low-High",  "Low-Low",
                                   "Not Significant"))
  )

preds       <- unique(predictor_lisa_classified$predictor)
pred_groups <- split(preds, ceiling(seq_along(preds) / 2))

for (grp in pred_groups) {
 maps   <- lapply(grp, function(pred) {
    predictor_lisa_map_data |>
      filter(predictor == pred) |>
      lisa_map(paste(pred, "LISA Clusters"))
  })
  second <- if (length(maps) == 2) maps[[2]] else blank_panel
  print(tmap_arrange(maps[[1]], second, ncol = 2, outer.margins = 0))
}
```

\newpage

# Ordinary Least Squares (OLS)

OLS provides a global baseline that assumes each predictor's relationship with crime is constant across all tracts. All predictors are standardized to mean 0, SD 1 so that coefficients reflect the change in log1p crime rate per one standard deviation increase in the predictor.

```{r ols}
ols_formula <- function(outcome) {
  as.formula(paste(outcome, "~", paste(model_vars, collapse = " + ")))
}

ols_models <- map(names(crime_rate_labels), function(var) {
  lm(ols_formula(var), data = st_drop_geometry(tract_data) |>
       mutate(across(all_of(model_vars), ~ as.numeric(scale(.x)))))
}) |>
  set_names(names(crime_rate_labels))
```

```{r ols-summary-violent}
tidy(ols_models[["log1p_rate_violent"]]) |>
  mutate(
    term = recode(term,
      "(Intercept)"        = "Intercept",
      !!!setNames(acs_var_labels, acs_model_vars)
    ),
    sig = case_when(
      p.value < 0.001 ~ "***",
      p.value < 0.01  ~ "**",
      p.value < 0.05  ~ "*",
      TRUE            ~ ""
    )
  ) |>
  select(term, estimate, std.error, statistic, p.value, sig) |>
  knitr::kable(
    caption   = "OLS Coefficients: Violent Crime",
    digits    = 3,
    col.names = c("Predictor", "Estimate", "Std. Error", "t", "p-value", ""),
    align     = c("l", "r", "r", "r", "r", "l")
  ) |>
  kable_style()
```

```{r ols-summary-property}
tidy(ols_models[["log1p_rate_property"]]) |>
  mutate(
    term = recode(term,
      "(Intercept)"        = "Intercept",
      !!!setNames(acs_var_labels, acs_model_vars)
    ),
    sig = case_when(
      p.value < 0.001 ~ "***",
      p.value < 0.01  ~ "**",
      p.value < 0.05  ~ "*",
      TRUE            ~ ""
    )
  ) |>
  select(term, estimate, std.error, statistic, p.value, sig) |>
  knitr::kable(
    caption   = "OLS Coefficients: Property Crime",
    digits    = 3,
    col.names = c("Predictor", "Estimate", "Std. Error", "t", "p-value", ""),
    align     = c("l", "r", "r", "r", "r", "l")
  ) |>
  kable_style()
```

\newpage 

## Model Fit Assessment

```{r ols-fit}
map_dfr(names(crime_rate_labels), function(var) {
  fit <- ols_models[[var]]
  glance(fit) |>
    mutate(
      Outcome = crime_rate_labels[var],
      AICc    = round(ols_aicc(fit), 3)
    ) |>
    select(Outcome, r.squared, adj.r.squared, AICc, BIC, df.residual)
}) |>
  knitr::kable(
    caption = "OLS Model Fit Statistics",
    digits  = 3,
    align   = c("l", "r", "r", "r", "r", "r")
  ) |>
  kable_style()
```

```{r ols-residuals-morans}
map_dfr(names(crime_rate_labels), function(var) {
  resid  <- residuals(ols_models[[var]])
  result <- moran.mc(resid, listw = w_queen, nsim = 999, zero.policy = TRUE)
  tibble(
    Outcome     = crime_rate_labels[var],
    `Moran's I` = round(result$statistic, 4),
    `p-value`   = round(result$p.value,   4)
  )
}) |>
  knitr::kable(
    caption = "Moran's I on OLS Residuals",
    align   = c("l", "r", "r")
  ) |>
  kable_style()
```

```{r ols-residuals-extract}
ols_resid_data <- map_dfr(names(crime_rate_labels), function(var) {
  tract_data |>
    st_drop_geometry() |>
    select(GEOID) |>
    mutate(
      outcome  = crime_rate_labels[var],
      residual = residuals(ols_models[[var]])
    )
})
```

\newpage

# Geographically Weighted Regression (GWR)

GWR fits a separate weighted regression at each tract where nearby tracts contribute more to local estimates than distant ones.

## Bandwidth Selection

```{r bandwidth-selection, cache=TRUE}
centroids <- st_coordinates(st_centroid(tract_data))
tract_sp <- tract_data |>
  mutate(
    across(all_of(model_vars), ~ as.numeric(scale(.x))),
    X = centroids[, 1],
    Y = centroids[, 2]
  ) |>
  st_drop_geometry() |>
  as.data.frame()

coordinates(tract_sp) <- ~ X + Y

invisible(capture.output({
  bw_violent <- bw.gwr(
    formula  = as.formula(paste("log1p_rate_violent ~", paste(model_vars, collapse = " + "))),
    data     = tract_sp,
    approach = "AICc",
    kernel   = "bisquare",
    adaptive = TRUE
  )

  bw_property <- bw.gwr(
    formula  = as.formula(paste("log1p_rate_property ~", paste(model_vars, collapse = " + "))),
    data     = tract_sp,
    approach = "AICc",
    kernel   = "bisquare",
    adaptive = TRUE
  )
}))

tibble(
  Outcome   = c("Violent Crime", "Property Crime"),
  Bandwidth = c(bw_violent, bw_property),
  Type      = "Adaptive (Nearest Neighbors)",
  Kernel    = "Bisquare",
  Method    = "AICc"
) |>
  knitr::kable(
    caption = "GWR Bandwidth Selection Results",
    align   = c("l", "r", "l", "l", "l")
  ) |>
  kable_style()
```

## Local Variance Inflation Factor (LVIF)

```{r lvif, cache=TRUE}
invisible(capture.output({
  lvif_violent <- gwr.collin.diagno(
    formula  = as.formula(paste("log1p_rate_violent ~", paste(model_vars, collapse = " + "))),
    data     = tract_sp,
    bw       = bw_violent,
    kernel   = "bisquare",
    adaptive = TRUE
  )

  lvif_property <- gwr.collin.diagno(
    formula  = as.formula(paste("log1p_rate_property ~", paste(model_vars, collapse = " + "))),
    data     = tract_sp,
    bw       = bw_property,
    kernel   = "bisquare",
    adaptive = TRUE
  )
}))
```

```{r lvif-summary}
summarise_lvif <- function(data) {
  data |>
    select(ends_with("_VIF")) |>
    summarise(across(everything(), list(
      Mean = \(x) round(mean(x, na.rm = TRUE), 3),
      Max  = \(x) round(max(x,  na.rm = TRUE), 3)
    ))) |>
    pivot_longer(
      everything(),
      names_to      = c("Predictor", ".value"),
      names_pattern = "(.+)_VIF_(Mean|Max)"
    ) |>
    mutate(Predictor = acs_var_labels[Predictor])
}

summarise_lvif(lvif_violent$SDF@data) |>
  rename(Violent_Mean = Mean, Violent_Max = Max) |>
  left_join(
    summarise_lvif(lvif_property$SDF@data) |>
      rename(Property_Mean = Mean, Property_Max = Max),
    by = "Predictor"
  ) |>
  arrange(desc(Violent_Max)) |>
  knitr::kable(
    caption   = "LVIF Summary: Violent and Property Crime",
    align     = c("l", "r", "r", "r", "r"),
    col.names = c("Predictor", "Mean", "Max", "Mean", "Max")
  ) |>
  kable_style() |>
  add_header_above(c(" " = 1, "Violent" = 2, "Property" = 2))
```

\newpage

## Model Fit Comparison

```{r gwr, cache=TRUE}
invisible(capture.output({
  gwr_violent <- gwr.basic(
    formula  = as.formula(paste("log1p_rate_violent ~", paste(model_vars, collapse = " + "))),
    data     = tract_sp,
    bw       = bw_violent,
    kernel   = "bisquare",
    adaptive = TRUE
  )

  gwr_property <- gwr.basic(
    formula  = as.formula(paste("log1p_rate_property ~", paste(model_vars, collapse = " + "))),
    data     = tract_sp,
    bw       = bw_property,
    kernel   = "bisquare",
    adaptive = TRUE
  )
}))
```

```{r gwr-fit}
tibble(
  Outcome          = c("Violent Crime", "Property Crime"),
  `OLS R²`         = c(round(summary(ols_models[["log1p_rate_violent"]])$r.squared,  3),
                       round(summary(ols_models[["log1p_rate_property"]])$r.squared, 3)),
  `GWR R²`         = c(round(gwr_violent$GW.diagnostic$gw.R2,  3),
                       round(gwr_property$GW.diagnostic$gw.R2, 3)),
  `R² Improvement` = c(round(gwr_violent$GW.diagnostic$gw.R2  
                       - summary(ols_models[["log1p_rate_violent"]])$r.squared,  3),
                       round(gwr_property$GW.diagnostic$gw.R2 
                       - summary(ols_models[["log1p_rate_property"]])$r.squared, 3)),
  `OLS AICc`       = c(round(ols_aicc(ols_models[["log1p_rate_violent"]]),  1),
                        round(ols_aicc(ols_models[["log1p_rate_property"]]), 1)),
  `GWR AICc`       = c(round(gwr_violent$GW.diagnostic$AICc,  1),
                        round(gwr_property$GW.diagnostic$AICc, 1))
  ) |>
  knitr::kable(
    caption = "OLS vs GWR Model Fit Comparison",
    align   = c("l", "r", "r", "r", "r", "r")
  ) |>
  kable_style()
```

```{r gwr-residuals-morans}
map_dfr(list(
  list(model = gwr_violent,  label = "Violent Crime"),
  list(model = gwr_property, label = "Property Crime")
), function(x) {
  resid  <- x$model$SDF$residual
  result <- moran.mc(resid, listw = w_queen, nsim = 999, zero.policy = TRUE)
  tibble(
    Outcome     = x$label,
    `Moran's I` = round(result$statistic, 4),
    `p-value`   = round(result$p.value,   4)
  )
}) |>
  knitr::kable(
    caption = "Moran's I on GWR Residuals",
    align   = c("l", "r", "r")
  ) |>
  kable_style()
```

```{r gwr-residuals-extract}
gwr_resid_data <- map_dfr(list(
  list(model = gwr_violent,  label = "Violent Crime"),
  list(model = gwr_property, label = "Property Crime")
), function(x) {
  tibble(
    GEOID    = tract_data$GEOID,
    outcome  = x$label,
    residual = x$model$SDF$residual
  )
})
```

```{r gwr-r2-extract}
gwr_r2_data <- map_dfr(list(
  list(model = gwr_violent,  label = "Violent Crime"),
  list(model = gwr_property, label = "Property Crime")
), function(x) {
  tract_data |>
    st_drop_geometry() |>
    select(GEOID) |>
    mutate(
      outcome  = x$label,
      local_r2 = x$model$SDF$Local_R2
    )
})
```

\newpage

# Multiscale Geographically Weighted Regression (MGWR)

MGWR extends GWR by allowing each predictor to operate at its own spatial scale, estimating a separate bandwidth per variable.

## Bandwidth Selection

```{r mgwr, cache=TRUE}
invisible(capture.output({
  mgwr_violent <- gwr.multiscale(
    formula   = as.formula(paste("log1p_rate_violent ~", paste(model_vars, collapse = " + "))),
    data      = tract_sp,
    adaptive  = TRUE,
    kernel    = "bisquare",
    criterion = "CVR",
    bws0      = rep(bw_violent, length(model_vars) + 1),
    verbose   = FALSE
  )

  mgwr_property <- gwr.multiscale(
    formula   = as.formula(paste("log1p_rate_property ~", paste(model_vars, collapse = " + "))),
    data      = tract_sp,
    adaptive  = TRUE,
    kernel    = "bisquare",
    criterion = "CVR",
    bws0      = rep(bw_property, length(model_vars) + 1),
    verbose   = FALSE
  )
}))
```

```{r mgwr-bandwidths}
tibble(
  Predictor        = acs_var_labels[model_vars],
  `Violent Crime`  = mgwr_violent$GW.arguments$bws[-1],
  `Property Crime` = mgwr_property$GW.arguments$bws[-1]
) |>
  arrange(`Violent Crime`) |>
  knitr::kable(
    caption = "MGWR Per-Predictor Bandwidths",
    align   = c("l", "r", "r")
  ) |>
  kable_style()
```

## Model Fit Comparison

```{r mgwr-fit}
tibble(
  Outcome        = c("Violent Crime", "Property Crime"),
  `OLS R²`  = c(round(summary(ols_models[["log1p_rate_violent"]])$r.squared,  3),
                      round(summary(ols_models[["log1p_rate_property"]])$r.squared, 3)),
  `GWR R²`  = c(round(gwr_violent$GW.diagnostic$gw.R2,  3),
                      round(gwr_property$GW.diagnostic$gw.R2, 3)),
  `MGWR R²` = c(round(mgwr_violent$GW.diagnostic$R2.val,  3),
                      round(mgwr_property$GW.diagnostic$R2.val, 3)),
  `OLS AICc`     = c(round(ols_aicc(ols_models[["log1p_rate_violent"]]),  1),
                      round(ols_aicc(ols_models[["log1p_rate_property"]]), 1)),
  `GWR AICc`     = c(round(gwr_violent$GW.diagnostic$AICc,  1),
                      round(gwr_property$GW.diagnostic$AICc, 1)),
  `MGWR AICc`    = c(round(mgwr_violent$GW.diagnostic$AICc,  1),
                      round(mgwr_property$GW.diagnostic$AICc, 1))
) |>
  knitr::kable(
    caption = "OLS vs GWR vs MGWR Model Fit Comparison",
    align   = c("l", "r", "r", "r", "r", "r", "r")
  ) |>
  kable_style()
```

```{r mgwr-residuals-morans}
map_dfr(list(
  list(model = mgwr_violent,  label = "Violent Crime"),
  list(model = mgwr_property, label = "Property Crime")
), function(x) {
  resid  <- x$model$SDF$residual
  result <- moran.mc(resid, listw = w_queen, nsim = 999, zero.policy = TRUE)
  tibble(
    Outcome     = x$label,
    `Moran's I` = round(result$statistic, 4),
    `p-value`   = round(result$p.value,   4)
  )
}) |>
  knitr::kable(
    caption = "Moran's I on MGWR Residuals",
    align   = c("l", "r", "r")
  ) |>
  kable_style()
```

## Residuals Maps

```{r mgwr-residuals-extract}
mgwr_resid_data <- map_dfr(list(
  list(model = mgwr_violent,  label = "Violent Crime"),
  list(model = mgwr_property, label = "Property Crime")
), function(x) {
  tibble(
    GEOID    = tract_data$GEOID,
    outcome  = x$label,
    residual = x$model$SDF$residual
  )
})
```

\newpage

## Coefficient Overview

```{r mgwr-coef-summary}
mgwr_summary_data <- bind_rows(
  mgwr_coef_summary(mgwr_violent,  "Violent Crime"),
  mgwr_coef_summary(mgwr_property, "Property Crime")
)

##identify predictors with at least one significant tract per outcome
mgwr_sig_violent  <- mgwr_summary_data |>
  filter(Outcome == "Violent Crime",  pct_sig_num > 0) |>
  pull(Predictor)

mgwr_sig_property <- mgwr_summary_data |>
  filter(Outcome == "Property Crime", pct_sig_num > 0) |>
  pull(Predictor)

mgwr_summary_data |>
  select(-Outcome, -pct_sig_num) |>
  knitr::kable(
    caption = "MGWR Coefficient Summary by Outcome",
    align   = c("l", "r", "r", "r", "r", "r")
  ) |>
  kable_style() |>
  kableExtra::pack_rows("Violent Crime",  1, length(model_vars)) |>
  kableExtra::pack_rows("Property Crime", length(model_vars) + 1, length(model_vars) * 2)
```

```{r mgwr-coef-setup}
##extract significant mgwr coefficients
mgwr_violent_map_data  <- extract_coef_data(mgwr_violent$SDF@data,  "Violent Crime")
mgwr_property_map_data <- extract_coef_data(mgwr_property$SDF@data, "Property Crime")

##pool all significant coefficients across all predictors and both outcomes for a unified scale
all_sig_coefs   <- c(mgwr_violent_map_data$coef, mgwr_property_map_data$coef)
coef_brks_global <- sym_breaks(all_sig_coefs[!is.na(all_sig_coefs)])
```

\newpage

## Coefficient Maps

```{r mgwr-coef-maps-paired, fig.show='asis'}
mgwr_sig_either <- union(mgwr_sig_violent, mgwr_sig_property)

for (var in model_vars) {
  label <- acs_var_labels[var]
  if (!label %in% mgwr_sig_either) next

  v_map <- if (label %in% mgwr_sig_violent)
    coef_map(mgwr_violent_map_data,  label, "Violent Crime",
             paste("Violent Crime:", label),  coef_brks_global) else blank_panel
  p_map <- if (label %in% mgwr_sig_property)
    coef_map(mgwr_property_map_data, label, "Property Crime",
             paste("Property Crime:", label), coef_brks_global) else blank_panel

  print(tmap_arrange(v_map, p_map, ncol = 2, outer.margins = 0))
}
```

## Local R² Extraction

```{r mgwr-r2-extract}
if (isTRUE(st_is_longlat(tract_data))) {
  tract_proj <- st_transform(tract_data, 3310)
} else {
  tract_proj <- tract_data
}

coords   <- st_coordinates(st_centroid(tract_proj))
dist_mat <- as.matrix(dist(coords))

##minimum bandwidth per model
bw_k_violent  <- min(mgwr_violent$GW.arguments$bws[-1])
bw_k_property <- min(mgwr_property$GW.arguments$bws[-1])

mgwr_r2_data <- bind_rows(
  tract_data |>
    st_drop_geometry() |>
    select(GEOID) |>
    mutate(
      outcome  = "Violent Crime",
      local_r2 = mgwr_local_r2(mgwr_violent, "log1p_rate_violent",
                                bw_k_violent, dist_mat)
    ),
  tract_data |>
    st_drop_geometry() |>
    select(GEOID) |>
    mutate(
      outcome  = "Property Crime",
      local_r2 = mgwr_local_r2(mgwr_property, "log1p_rate_property",
                                bw_k_property, dist_mat)
    )
)
```

\newpage

## Residuals Maps

```{r global-scales}
##pool all residuals across OLS, GWR, and MGWR to define one shared symmetric scale
resid_brks_global <- sym_breaks(
  c(ols_resid_data$residual, gwr_resid_data$residual, mgwr_resid_data$residual)
)

##pool all local r2 values across GWR and MGWR to define one shared sequential scale
r2_brks_global <- seq_breaks(c(gwr_r2_data$local_r2, mgwr_r2_data$local_r2))
```

```{r resid-maps-ols}
tmap_arrange(
  resid_map(ols_resid_data, "Violent Crime",  "Violent Crime: OLS Residuals",  resid_brks_global),
  resid_map(ols_resid_data, "Property Crime", "Property Crime: OLS Residuals", resid_brks_global),
  ncol = 2, outer.margins = 0
)
```

```{r resid-maps-gwr}
tmap_arrange(
  resid_map(gwr_resid_data, "Violent Crime",  "Violent Crime: GWR Residuals",  resid_brks_global),
  resid_map(gwr_resid_data, "Property Crime", "Property Crime: GWR Residuals", resid_brks_global),
  ncol = 2, outer.margins = 0
)
```

```{r resid-maps-mgwr}
tmap_arrange(
  resid_map(mgwr_resid_data, "Violent Crime",  "Violent Crime: MGWR Residuals",  resid_brks_global),
  resid_map(mgwr_resid_data, "Property Crime", "Property Crime: MGWR Residuals", resid_brks_global),
  ncol = 2, outer.margins = 0
)
```

\newpage

## Local R² Maps

```{r r2-maps-gwr}
tmap_arrange(
  local_r2_map(gwr_r2_data, "Violent Crime",  "Violent Crime: GWR Local R²",  r2_brks_global),
  local_r2_map(gwr_r2_data, "Property Crime", "Property Crime: GWR Local R²", r2_brks_global),
  ncol = 2, outer.margins = 0
)
```

```{r r2-maps-mgwr}
tmap_arrange(
  local_r2_map(mgwr_r2_data, "Violent Crime",  "Violent Crime: MGWR Local R²",  r2_brks_global),
  local_r2_map(mgwr_r2_data, "Property Crime", "Property Crime: MGWR Local R²", r2_brks_global),
  ncol = 2, outer.margins = 0
)
```

\newpage

# Bivariate LISA

Bivariate LISA examines the spatial co-clustering of each significant MGWR predictor with crime rates, testing whether tracts high in a given socioeconomic indicator are surrounded by tracts with high crime (and vice versa).

```{r bv-lisa-compute}
tract_bv <- tract_data |>
  filter(GEOID != isolated_geoid)

nb_bv <- poly2nb(tract_bv, queen = TRUE)
w_bv  <- nb2listw(nb_bv, style = "W", zero.policy = FALSE)

bv_pair_df <- bind_rows(
  tibble(label_x = mgwr_sig_violent,  y = "log1p_rate_violent",  label_y = "Violent Crime"),
  tibble(label_x = mgwr_sig_property, y = "log1p_rate_property", label_y = "Property Crime")
) |>
  mutate(x = names(acs_var_labels)[match(label_x, acs_var_labels)]) |>
  arrange(
    match(x, model_vars),
    match(label_y, c("Violent Crime", "Property Crime"))
  )

bv_pairs <- pmap(bv_pair_df, function(x, y, label_x, label_y) {
  list(x = x, y = y, label_x = label_x, label_y = label_y)
})

bv_lisa_results <- map(bv_pairs, function(p) {
  x_std <- scale(tract_bv[[p$x]])[, 1]
  y_std <- scale(tract_bv[[p$y]])[, 1]

  lm_bv <- localmoran_bv(x_std, y_std, listw = w_bv, nsim = 999)

  lag_y <- lag.listw(w_bv, y_std)
  p_val <- lm_bv[, "Pr(z != E(Ibvi))"]

  tract_bv |>
    st_drop_geometry() |>
    select(GEOID) |>
    mutate(
      label   = paste(p$label_x, "x", p$label_y),
      label_x = p$label_x,
      label_y = p$label_y,
      bv_i    = lm_bv[, "Ibvi"],
      p_val   = p_val,
      x_std   = x_std,
      lag_y   = lag_y,
      quad    = case_when(
        p_val >= 0.01            ~ "Not Significant",
        x_std > 0 & lag_y > 0   ~ "High-High",
        x_std > 0 & lag_y <= 0  ~ "High-Low",
        x_std <= 0 & lag_y > 0  ~ "Low-High",
        x_std <= 0 & lag_y <= 0 ~ "Low-Low"
      )
    )
})

names(bv_lisa_results) <- map_chr(bv_pairs, ~ paste(.x$label_x, "x", .x$label_y))
```

```{r bv-lisa-table}
bv_summary <- map_dfr(bv_lisa_results, function(df) {
  tibble(
    label_x               = df$label_x[1],
    Outcome               = df$label_y[1],
    `Mean Local Moran's I` = round(mean(df$bv_i, na.rm = TRUE), 4),
    `High-High`           = sum(df$quad == "High-High"),
    `Low-Low`             = sum(df$quad == "Low-Low"),
    `% Significant`       = paste0(round(mean(df$p_val < 0.01) * 100, 1), "%")
  )
})

##compute pack_rows group boundaries
group_bounds <- bv_summary |>
  mutate(row = row_number()) |>
  group_by(label_x) |>
  summarise(start = min(row), end = max(row), .groups = "drop")

kbl <- bv_summary |>
  select(-label_x) |>
  knitr::kable(
    caption = "Bivariate LISA Summary: Significant MGWR Predictors x Crime Outcome",
    align   = c("l", "r", "r", "r", "r")
  ) |>
  kable_style()

for (i in seq_len(nrow(group_bounds))) {
  kbl <- kbl |>
    kableExtra::pack_rows(group_bounds$label_x[i], group_bounds$start[i], group_bounds$end[i])
}

kbl
```

## Maps

```{r bv-lisa-maps-paired, fig.show='asis'}
bv_pred_names <- unique(map_chr(bv_pairs, ~ .x$label_x))

for (pred in bv_pred_names) {
  v_nm <- paste(pred, "x Violent Crime")
  p_nm <- paste(pred, "x Property Crime")

  make_bv_map <- function(nm, legend_title) {
    tract_data |>
      left_join(
        bv_lisa_results[[nm]] |>
          select(GEOID, quad) |>
          mutate(quad = factor(quad, levels = c("High-High", "High-Low",
                                                "Low-High",  "Low-Low",
                                                "Not Significant"))),
        by = "GEOID"
      ) |>
      lisa_map(nm, legend_title)
  }

  v_map <- if (v_nm %in% names(bv_lisa_results)) make_bv_map(v_nm, "BV LISA Cluster") else blank_panel
  p_map <- if (p_nm %in% names(bv_lisa_results)) make_bv_map(p_nm, "BV LISA Cluster") else blank_panel

  print(tmap_arrange(v_map, p_map, ncol = 2, outer.margins = 0))
}
```
